---
title: "Commission on Higher Education"
subtitle: "Budget evolution and agency-level utilization, FY 2018–2026"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    self_contained: true
    df_print: paged
    fig_width: 10
    fig_height: 5.5
---

```{css, echo=FALSE}
:root { --ink:#1a1a1a; --muted:#6e6e6e; --accent:#08519C; --rule:#e6e6e6; }
body, .main-container { font-family: Arial, Helvetica, "IBM Plex Sans", sans-serif; color: var(--ink); }
.main-container { max-width: 980px; }
h1, h2, h3, h4 { font-family: Arial, Helvetica, sans-serif; font-weight: 600; }
h1.title { font-size: 30px; }
h1 { font-size: 25px; margin-top: 1.6em; padding-top: .3em; border-top: 2px solid var(--rule); }
h2 { font-size: 20px; margin-top: 1.4em; }
h3 { font-size: 17px; color: #333; }
.author, .date { color: var(--muted); }
p, li { font-size: 15.5px; line-height: 1.62; }
blockquote { border-left: 3px solid var(--accent); background: #f5f8fc; color: #25405e;
  font-size: 14.5px; padding: .6em 1em; margin: 1.2em 0; }
.lead { font-size: 16.5px; color: #2b2b2b; }
.takeaway { background: #f7f9fb; border: 1px solid var(--rule); border-radius: 6px;
  padding: .4em 1.1em; margin: 1.1em 0; }
.takeaway p, .takeaway li { font-size: 14.5px; }
.caption, .figcaption { color: var(--muted); font-size: 13px; }
table.dataTable { font-size: 13px; }
.kbl-note { color: var(--muted); font-size: 12.5px; margin-top: -.5em; }
hr { border-top: 1px solid var(--rule); }
.home-link { font-size: 13.5px; }
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                     dpi = 150, fig.align = "center", out.width = "100%")
options(knitr.kable.NA = "")
library(readxl); library(dplyr); library(tidyr); library(stringr)
library(ggplot2); library(scales); library(knitr); library(kableExtra); library(DT)

.candidates <- c("../data/Compiled_-_NEP-GAA-NoFAR1.xlsx", "data/Compiled_-_NEP-GAA-NoFAR1.xlsx",
                 "Compiled_-_NEP-GAA-NoFAR1.xlsx")
file_path <- .candidates[file.exists(.candidates)][1]
stopifnot(!is.na(file_path))
AGENCY_SHEET <- "CHED"; EXEC_SHEET <- "SAAODB"

theme_ched <- function(base_size = 12) {
  theme_minimal(base_size = base_size, base_family = "sans") +
    theme(
      plot.title = element_text(face = "bold", size = base_size + 2, color = "black"),
      plot.subtitle = element_text(color = "grey30", size = base_size),
      plot.caption = element_text(color = "grey40", size = base_size - 3, hjust = 0),
      panel.grid.minor = element_blank(), panel.grid.major.x = element_blank(),
      axis.title = element_text(size = base_size - 1, color = "grey20"),
      axis.text = element_text(color = "grey20"),
      legend.position = "bottom", legend.title = element_blank(),
      legend.text = element_text(size = base_size - 1),
      strip.text = element_text(face = "bold", size = base_size - 1.5, lineheight = 0.9))
}

exec_pal <- c("NEP (proposed)" = "#9E9AC8", "GAA (enacted)" = "#54278F",
  "Allotments" = "#08519C", "Obligations" = "#E6550D", "Disbursements" = "#31A354")
cat_pal <- c(
  "General Administration & Support"                = "#A6761D",
  "Support to Operations"                           = "#E6AB02",
  "Higher Education Regulation & Quality Assurance" = "#7570B3",
  "Higher Education Development & Student Subsidies" = "#1B9E77")
absorp_pal <- c("Obligations / Allotment" = "#3182BD",
  "Disbursements / Obligations" = "#E6550D", "Disbursements / Allotment" = "#31A354")
ec_pal <- c("Personnel Services (PS)" = "#2CA02C",
  "Maintenance & Other Op. Exp. (MOOE)" = "#FF7F0E", "Capital Outlays (CO)" = "#9467BD")
bar_fill <- "#08519C"; peso <- "PHP "
php_b      <- function(x) paste0(peso, comma(x/1e9, accuracy = 0.01), "B")
php_b_axis <- function(x) paste0(peso, comma(x/1e9, accuracy = 1), "B")

dt_table <- function(df, caption = NULL, page = 10, money_cols = NULL, pct_cols = NULL, money_digits = 2) {
  dt <- datatable(df, rownames = FALSE, caption = caption, extensions = "Buttons",
    options = list(pageLength = page, lengthMenu = c(10, 20, 50, 100), dom = "Bfrtip",
      buttons = c("copy", "csv"), scrollX = TRUE, autoWidth = FALSE,
      columnDefs = list(list(className = "dt-right", targets = which(sapply(df, is.numeric)) - 1))),
    class = "stripe hover compact")
  if (!is.null(money_cols)) { mc <- which(names(df) %in% money_cols)
    if (length(mc)) dt <- formatRound(dt, mc, digits = money_digits, mark = ",") }
  if (!is.null(pct_cols)) { pc <- which(names(df) %in% pct_cols)
    if (length(pc)) dt <- formatPercentage(dt, pc, digits = 0) }
  dt
}
kbl_clean <- function(x, font = 14, ...) {
  x %>% kable(...) %>% kable_styling(full_width = TRUE, font_size = font,
    bootstrap_options = c("condensed", "striped", "hover"))
}

# ---- Data load -----------------------------------------------------------
# CHED has been an Other Executive Office throughout the period (attached to the
# Office of the President), so there is a single departmental home and no
# reclassification to reconcile. The agency tab and the SAAODB tab may spell the
# agency slightly differently, so match on a normalised name.
CODE_COLS <- c("DEPARTMENT", "AGENCY", "PREXC_PROG", "PREXC_SUBPROG")
raw <- read_excel(file_path, sheet = AGENCY_SHEET) %>%
  mutate(across(any_of(CODE_COLS), as.character)) %>% filter(!is.na(`P/A/P`))
AGENCY_NAME <- raw$UACS_AGY_DSC[!is.na(raw$UACS_AGY_DSC)][1]
agency_key  <- function(x) str_trim(str_remove(x, "\\s*\\(.*\\)\\s*$"))

df_long_all <- raw %>%
  pivot_longer(cols = matches("^(NEP|GAA|AuthAppro|AdjAppro|AdjAllot|Obligations|Disbursements)_\\d{4}_EXP_(1PS|2MOOE|6CO|TOTAL)$"),
               names_to = "var", values_to = "amount") %>%
  separate(var, into = c("metric", "year", "tag", "expense_class"), sep = "_") %>%
  mutate(year = as.integer(year), amount = as.numeric(amount)) %>%
  select(-tag) %>% rename(agency = UACS_AGY_DSC, dept = UACS_DPT_DSC, pap = `P/A/P`)

df_long <- df_long_all %>% filter(expense_class == "TOTAL") %>% select(-expense_class) %>%
  mutate(category = case_when(
    PREXC_PROG == "1000" ~ "General Administration & Support",
    PREXC_PROG == "2000" ~ "Support to Operations",
    PREXC_PROG == "3101" ~ "Higher Education Regulation & Quality Assurance",
    PREXC_PROG == "3102" ~ "Higher Education Development & Student Subsidies",
    TRUE ~ NA_character_))

totals <- df_long %>% group_by(year, metric) %>%
  summarise(total = if (all(is.na(amount))) NA_real_ else sum(amount, na.rm = TRUE), .groups = "drop")

pap_year <- df_long %>% group_by(pap, year, metric) %>%
  summarise(amount = if (all(is.na(amount))) NA_real_ else sum(amount, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = metric, values_from = amount)

exec <- read_excel(file_path, sheet = EXEC_SHEET) %>%
  filter(agency_key(Agency) == agency_key(AGENCY_NAME)) %>%
  pivot_longer(cols = matches("^\\d{4}$"), names_to = "year", values_to = "amt_k") %>%
  mutate(year = as.integer(year), amount = amt_k * 1e3) %>% filter(!is.na(amount)) %>%
  group_by(year, Particular) %>% summarise(amount = sum(amount, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = Particular, values_from = amount) %>% rename(Obligations = `Obligations Incurred`) %>%
  mutate(oblig_allot = Obligations / Allotments, disb_oblig = Disbursements / Obligations,
         disb_allot = Disbursements / Allotments) %>% arrange(year)
```

<p class="home-link"><a href="index.html">← All agencies</a></p>

# Overview {#overview .unnumbered}

<p class="lead">
This document profiles the budget of the **Commission on Higher Education (CHED)** — the government's body for regulating and developing higher education — across nine fiscal years, FY 2018 to FY 2026. CHED is a **large, multi-billion-peso agency**, but it carries the same data constraint as several commissions: usable program-level execution records were not available, so this review pairs the **proposed-to-enacted budget at the program level** with **budget utilization measured at the agency aggregate**.
</p>

Because the program-by-program execution chain cannot be traced, the emphasis is on what CHED's budget is *composed of*, how Congress adjusts it, and — at the whole-agency level — how much of what is released actually gets obligated and paid out.

Three features define CHED's budget. First, it is, in budget terms, a **free-tuition fund**: a single program — student subsidies, dominated by **Universal Access to Quality Tertiary Education (UAQTEA)** — is about **96% of the money**, while CHED's own *regulatory* mandate is a rounding error. Second, the budget has **not grown**: it was ₱49.4B in 2018 and ₱47.5B in 2026, essentially flat across nine years. Third, it is **extraordinarily volatile** — Congress has swung the enacted budget from **+₱37B above the proposal (2018) to −₱20B below it (2022)** — and a persistent gap between what is released and what is spent makes weak execution its defining operational problem.

# Executive summary {#exec}

**CHED is a large but stagnant budget.** Enacted funding was **₱49.4B in 2018 and ₱47.5B in 2026** — essentially flat over nine years, having peaked around ₱50B (2018–2021) and bottomed near ₱31B (2022–2023).

**In budget terms, CHED is a free-tuition fund.** Student-subsidy programs are about **96% of the 2026 budget**, and a single P/A/P — **Universal Access to Quality Tertiary Education (UAQTEA)** — is **₱37.6B, about 79% of the whole agency** on its own.

**Its regulatory mandate is a rounding error.** Quality assurance and regulation — CHED's core statutory job — is about **2% of the budget** (~₱1.1B), with administration and support another 2%.

**Congress reshapes CHED violently.** It added **+₱37B in 2018** (when free higher education was first funded, against a ₱12B proposal), **cut ₱20B in 2022**, and added **+₱14B in 2026** — swings far larger than most agencies' entire budgets.

**Execution is strikingly weak.** Cash utilization (disbursements-to-allotment) is often just **40–60%**; large unspent balances build up on the free-tuition line year after year.

> CHED is, on paper, a regulator; in budget terms, it is a machine for disbursing free tuition — one that is not growing, swings wildly with each budget cycle, and pays out well under two-thirds of what it is given. The pressing questions are why the money is not reaching students, and whether a flat budget can sustain a demand-driven entitlement.

# A note on the data {#data-notes}

**Period.** FY 2018 through FY 2026 for the proposed (NEP) and enacted (GAA) budget, at the P/A/P level. Agency-level execution (Allotments, Obligations, Disbursements) covers FY 2011–2025.

**No program-level execution.** Usable **FAR No. 1** (SAAODB) records were not available for CHED, so there is **no P/A/P-level absorption analysis**. Utilization is shown only at the **agency aggregate** — the gap this review would most like to close, given how central the execution question is for CHED.

**Two bases, not directly comparable.** The NEP/GAA figures are **current-year new appropriations only**. The execution figures are DBM's **published agency aggregates, which combine all available funds** — new, automatic, and continuing appropriations. This distinction matters especially for CHED, because unspent free-tuition balances carry forward as continuing appropriations. The two series should not be read one-to-one.

**A single home.** CHED has been an **Other Executive Office** (attached to the Office of the President) throughout the period, so there is no reclassification to reconcile and its four-program structure is stable — the budget series is directly comparable across all years.

**Expense classes.** Personnel Services (PS), Maintenance & Other Operating Expenses (MOOE), Capital Outlays (CO). **Units are PHP billions** throughout.

# The aggregate budget {#aggregate}

## Large, flat, and volatile

```{r evolution, fig.height=5}
evo <- totals %>% filter(metric %in% c("NEP","GAA")) %>%
  mutate(metric = factor(metric, levels = c("NEP","GAA"), labels = c("NEP (proposed)","GAA (enacted)")))
ggplot(evo, aes(year, total, color = metric, group = metric)) +
  geom_line(linewidth = 1.0) + geom_point(size = 2) +
  scale_color_manual(values = exec_pal) +
  scale_y_continuous(labels = php_b_axis, breaks = pretty_breaks(6), limits = c(0, NA), expand = expansion(mult = c(0, .05))) +
  scale_x_continuous(breaks = seq(2018, 2026, 2)) +
  labs(title = "CHED current-year appropriations, FY 2018–2026",
       subtitle = "No net growth in nine years (~₱47–49B), but a violent 2022 trough and a wide proposal-to-enacted gap",
       x = NULL, y = NULL, caption = "Current-year new appropriations. Source: DBM NEP & GAA.") +
  theme_ched() + guides(color = guide_legend(nrow = 1))
```

Two things stand out. There is **no net growth**: CHED is about ₱47B in both 2018 and 2026, having peaked near ₱50B (2018–2021) and dropped to about ₱31B in 2022–2023. And the enacted (GAA) line **swings hugely around the proposal (NEP)** — most dramatically in 2018, when the proposal was only ₱12B but the enacted budget was ₱49B, and in 2022, when a ₱52B proposal was cut to ₱32B. CHED's budget is as much a product of the *legislature* as of the Executive.

```{r tbl-evolution}
totals %>% filter(metric %in% c("NEP","GAA")) %>% mutate(value = total/1e9) %>%
  select(year, metric, value) %>% pivot_wider(names_from = year, values_from = value) %>%
  mutate(metric = factor(metric, levels = c("NEP","GAA"))) %>% arrange(metric) %>% rename(`PHP B` = metric) %>%
  kbl_clean(font = 13, digits = 1, format.args = list(big.mark = ","), na = "")
```

## Where Congress adjusts the budget

```{r nep-gaa-gap, fig.height=4.4}
gap <- totals %>% filter(metric %in% c("NEP","GAA")) %>% pivot_wider(names_from = metric, values_from = total) %>%
  mutate(change = GAA - NEP, direction = ifelse(change >= 0, "Net increase (GAA > NEP)", "Net decrease (GAA < NEP)"))
ggplot(gap, aes(year, change/1e9, fill = direction)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%+.1f", change/1e9), vjust = ifelse(change >= 0, -0.4, 1.2)), size = 3.1, color = "grey20") +
  scale_fill_manual(values = c("Net increase (GAA > NEP)" = "#1B9E77", "Net decrease (GAA < NEP)" = "#D95F02")) +
  scale_x_continuous(breaks = 2018:2026) +
  scale_y_continuous(labels = function(v) paste0(peso, v, "B"), expand = expansion(mult = c(0.14, 0.16))) +
  labs(title = "Enacted minus proposed (GAA − NEP), PHP billions",
       subtitle = "Enormous swings: +₱37B in 2018 (free HE first funded), −₱20B in 2022, +₱14B in 2026",
       x = NULL, y = NULL) +
  theme_ched() + guides(fill = guide_legend(nrow = 1))
```

The scale of these adjustments is unusual. The **+₱37B in 2018** reflects the first funding of free higher education — the Executive proposed ₱12B, and Congress raised the enacted budget to ₱49B. The **−₱20B in 2022** was the reverse: a ₱52B proposal cut to ₱32B. For a multi-year entitlement like free tuition, swings of this size are extremely hard to plan around.

## Which P/A/Ps does Congress augment?

```{r aug, fig.height=4.8}
aug <- df_long %>% filter(metric %in% c("NEP","GAA")) %>%
  group_by(pap, year, metric) %>% summarise(amount = sum(amount, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = metric, values_from = amount, values_fill = 0) %>%
  mutate(delta = GAA - NEP) %>% group_by(pap) %>% summarise(delta = sum(delta), .groups = "drop")
aug_plot <- bind_rows(aug %>% slice_max(delta, n = 6), aug %>% slice_min(delta, n = 4)) %>%
  filter(abs(delta) > 1e8) %>%
  mutate(dir = ifelse(delta >= 0, "Increased by Congress", "Cut by Congress"),
         pap = factor(pap, levels = pap[order(delta)]))
ggplot(aug_plot, aes(delta/1e9, pap, fill = dir)) +
  geom_col(width = 0.72) +
  geom_text(aes(label = sprintf("%+.1f", delta/1e9), hjust = ifelse(delta >= 0, -0.12, 1.12)), size = 3, color = "grey20") +
  scale_fill_manual(values = c("Increased by Congress" = "#1B9E77", "Cut by Congress" = "#D95F02")) +
  scale_x_continuous(labels = function(v) paste0(peso, v, "B"), expand = expansion(mult = c(0.18, 0.18))) +
  scale_y_discrete(labels = function(x) str_wrap(str_trunc(x, 54), 34)) +
  labs(title = "P/A/Ps most changed by Congress, cumulative FY 2018–2026 (GAA − NEP)",
       subtitle = "The free-tuition line (UAQTEA) absorbs nearly all of both the additions and the cuts",
       x = NULL, y = NULL) +
  theme_ched() + theme(panel.grid.major.y = element_blank(), panel.grid.major.x = element_line(color = "grey85"),
        axis.text.y = element_text(size = 9, lineheight = 0.9)) +
  guides(fill = guide_legend(nrow = 1))
```

Because the free-tuition line is so much larger than everything else, it is also where nearly all of Congress's adjustments land — both the big additions and the 2022 cut flow through **UAQTEA**. The smaller scholarship, subsidy, and regulatory lines are adjusted only at the margins.

# Where the money goes {#composition}

## Composition by PREXC program

```{r composition, fig.height=5}
comp <- df_long %>% filter(metric == "GAA", year <= 2026, !is.na(amount), !is.na(category)) %>%
  group_by(year, category) %>% summarise(amount = sum(amount, na.rm = TRUE), .groups = "drop") %>%
  mutate(category = factor(category, levels = names(cat_pal)))
ggplot(comp, aes(year, amount, fill = category)) +
  geom_col(width = 0.75, color = "white", linewidth = 0.2) +
  scale_fill_manual(values = cat_pal, labels = function(x) str_wrap(x, 32)) +
  scale_y_continuous(labels = php_b_axis, breaks = pretty_breaks(6), limits = c(0, NA), expand = expansion(mult = c(0, .05))) +
  scale_x_continuous(breaks = seq(2018, 2026, 2)) +
  labs(title = "Composition of the GAA by PREXC program, 2018–2026",
       subtitle = "Student subsidies are ~96% of the budget; regulation and administration are a thin band",
       x = NULL, y = NULL) +
  theme_ched() + guides(fill = guide_legend(nrow = 2, byrow = TRUE)) +
  theme(legend.text = element_text(size = 9), legend.key.size = unit(0.4, "cm"))
```

The green **Higher Education Development & Student Subsidies** program is essentially the whole budget — about 96% in 2026. What CHED is best known for as an institution — **regulation and quality assurance** (purple) — is a sliver, alongside administration and support. In budget terms, CHED's regulatory function and its subsidy function are almost entirely different in scale: the agency spends the overwhelming majority of its money moving tuition subsidies to students, not regulating institutions.

## Composition by expense class

```{r comp-ec, fig.height=5}
ec_levels <- c("Personnel Services (PS)","Maintenance & Other Op. Exp. (MOOE)","Capital Outlays (CO)")
comp_ec <- df_long_all %>% filter(metric == "GAA", year <= 2026, expense_class %in% c("1PS","2MOOE","6CO")) %>%
  group_by(year, expense_class) %>% summarise(amount = sum(amount, na.rm = TRUE), .groups = "drop") %>%
  mutate(expense_class = factor(expense_class, levels = c("1PS","2MOOE","6CO"), labels = ec_levels))
ggplot(comp_ec, aes(year, amount, fill = expense_class)) +
  geom_col(width = 0.75, color = "white", linewidth = 0.2) +
  scale_fill_manual(values = ec_pal) +
  scale_y_continuous(labels = php_b_axis, breaks = pretty_breaks(6), limits = c(0, NA), expand = expansion(mult = c(0, .05))) +
  scale_x_continuous(breaks = seq(2018, 2026, 2)) +
  labs(title = "GAA composition by expense class, 2018–2026",
       subtitle = "Almost entirely MOOE — tuition subsidies and grants classify as operating expenses",
       x = NULL, y = NULL) +
  theme_ched() + guides(fill = guide_legend(nrow = 1))
```

CHED is an overwhelmingly **MOOE** agency: because tuition subsidies and grants book as operating expenses, maintenance-and-operating spending dominates, with a thin personnel line and little capital. The expense-class mix simply mirrors the program mix — this is a budget for transferring money to and on behalf of students.

## The largest P/A/Ps

```{r top-paps, fig.height=4.8}
top_paps <- pap_year %>% filter(year == 2026, !is.na(GAA), GAA > 0) %>% arrange(desc(GAA)) %>%
  slice_head(n = 8) %>% mutate(pap = factor(pap, levels = rev(pap)))
ggplot(top_paps, aes(pap, GAA)) +
  geom_col(fill = bar_fill, width = 0.72) +
  geom_text(aes(label = php_b(GAA)), hjust = -0.1, size = 3.2, color = "grey20") +
  coord_flip() + scale_x_discrete(labels = function(x) str_wrap(x, 50)) +
  scale_y_continuous(labels = php_b_axis, expand = expansion(mult = c(0, .22)), limits = c(0, NA)) +
  labs(title = "Largest P/A/Ps by FY 2026 GAA",
       subtitle = "Universal Access to Quality Tertiary Education (₱37.6B) is about 79% of the agency on its own",
       x = NULL, y = NULL) +
  theme_ched() + theme(panel.grid.major.y = element_blank(), panel.grid.major.x = element_line(color = "grey85"))
```

The concentration is extreme: **UAQTEA (₱37.6B)** — free tuition at state universities and colleges — is nearly four-fifths of the entire agency, dwarfing the next lines: the **Tulong Dunong** grant program (₱2.7B), other scholarships and incentives (₱2.2B), post-graduate and medical scholarships (~₱1B each). CHED's regulatory and monitoring lines do not appear until far down the list.

### Full P/A/P ranking, FY 2026 (interactive) {.unnumbered}

```{r tbl-allpaps}
pap_year %>% filter(year == 2026, !is.na(GAA)) %>%
  left_join(pap_year %>% filter(year == 2025) %>% transmute(pap, GAA25 = GAA), by = "pap") %>%
  transmute(`P/A/P` = pap, `NEP 2026 (B)` = NEP/1e9, `GAA 2026 (B)` = GAA/1e9,
            `GAA 2025 (B)` = GAA25/1e9, `Net add 2026 (B)` = (GAA - NEP)/1e9) %>%
  arrange(desc(`GAA 2026 (B)`)) %>%
  dt_table(page = 10, money_cols = c("NEP 2026 (B)","GAA 2026 (B)","GAA 2025 (B)","Net add 2026 (B)"))
```

# Overall budget utilization {#utilization}

With no program-level execution data, utilization can only be read at the **agency aggregate**, and on a different basis from the appropriations above (all available funds, not just current-year new appropriations). On that basis CHED's execution is **strikingly weak** — a chronic gap between money released and money spent.

## Execution over time

```{r exec-levels, fig.height=4.8}
el <- exec %>% select(year, Allotments, Obligations, Disbursements) %>%
  pivot_longer(-year, names_to = "stage", values_to = "amount") %>% filter(!is.na(amount)) %>%
  mutate(stage = factor(stage, levels = c("Allotments","Obligations","Disbursements")))
ggplot(el, aes(year, amount, color = stage, group = stage)) +
  geom_line(linewidth = 1.0) + geom_point(size = 2) +
  scale_color_manual(values = exec_pal) +
  scale_y_continuous(labels = php_b_axis, breaks = pretty_breaks(6), limits = c(0, NA), expand = expansion(mult = c(0, .05))) +
  scale_x_continuous(breaks = seq(2011, 2025, 2)) +
  labs(title = "Budget execution, FY 2011–2025 (all available funds)",
       subtitle = "A wide, persistent gap between allotments and disbursements — much of the money is not paid out in-year",
       x = NULL, y = NULL, caption = "Includes new, automatic & continuing appropriations; not comparable to the NEP/GAA series.") +
  theme_ched() + guides(color = guide_legend(nrow = 1))
```

The gap between the **allotment** line and the **disbursement** line is the story: large sums are released but not spent within the year, and they accumulate as unspent balances that carry forward. For a program meant to pay tuition on students' behalf each semester, that persistent lag is the central concern.

## Absorptive capacity

```{r absorp, fig.height=4.8}
absorp_long <- exec %>% select(year, oblig_allot, disb_oblig, disb_allot) %>%
  pivot_longer(-year, names_to = "ratio", values_to = "value") %>% filter(!is.na(value)) %>%
  mutate(ratio = factor(ratio, levels = c("oblig_allot","disb_oblig","disb_allot"),
    labels = c("Obligations / Allotment","Disbursements / Obligations","Disbursements / Allotment")))
ggplot(absorp_long, aes(year, value, color = ratio, group = ratio)) +
  geom_line(linewidth = 0.9) + geom_point(size = 2) +
  scale_color_manual(values = absorp_pal) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0.3, 1.0), breaks = seq(0.3, 1, 0.1)) +
  scale_x_continuous(breaks = seq(2011, 2025, 2)) +
  labs(title = "CHED absorptive capacity, FY 2011–2025",
       subtitle = "Cash utilization runs low and erratic — disbursements-to-allotment is often just 40–60%",
       x = NULL, y = NULL) +
  theme_ched() + guides(color = guide_legend(nrow = 1))
```

CHED's absorption is both **low and erratic**. Disbursement-to-allotment has ranged from the high-30s to the high-60s percent, sitting in the **40–60%** band in most years — meaning that in a typical year, roughly half of what is released is not actually disbursed. Obligation rates are steadier but still volatile. Whatever the causes — eligibility verification, enrolment timing, coordination with state universities, or reporting lags — the result is that a large, flat budget is delivering considerably less to students than its headline suggests.

### Execution at a glance (interactive) {.unnumbered}

```{r exec-table}
exec %>% transmute(Year = year, `Allotments (B)` = Allotments/1e9, `Obligations (B)` = Obligations/1e9,
            `Disbursements (B)` = Disbursements/1e9, `O/A` = oblig_allot, `D/O` = disb_oblig, `D/A` = disb_allot) %>%
  dt_table(page = 15, money_cols = c("Allotments (B)","Obligations (B)","Disbursements (B)"),
           pct_cols = c("O/A","D/O","D/A"))
```

<p class="kbl-note">PHP billions, all-funds basis. O/A = obligations ÷ allotment; D/O = disbursements ÷ obligations; D/A = disbursements ÷ allotment. Not comparable to the NEP/GAA appropriations series above.</p>

# Questions for discussion {#discussion}

1. **A regulator that mostly disburses subsidies.** About 96% of CHED's budget is student subsidies; quality assurance and regulation — its statutory core — is around 2%. Is the balance between funding higher education and *governing* it the right one?
2. **Why isn't the money reaching students?** Only ~40–60% of the allotment is disbursed in-year, and unspent free-tuition balances build up. What is causing the lag — eligibility processing, enrolment timing, coordination with SUCs — and how much reaches students late or not at all?
3. **Flat funding, rising demand.** The budget has not grown since 2018 while enrolment and costs have risen. Can a flat appropriation sustain a demand-driven entitlement like free higher education, and what gives if it cannot?
4. **Living with volatility.** A +₱37B / −₱20B swing between years is hard to plan around for a multi-year commitment. How should a tuition-subsidy program be budgeted so that funding is predictable for students and institutions?
5. **What would program-level execution data show?** With no FAR No. 1, we can see that CHED's overall disbursement lags badly, but not *which* subsidy or scholarship lines are stalling. Program-level records would show exactly where the free-tuition money is getting stuck.

---

<p class="caption">
Underlying data: DBM NEP & GAA documents (current-year new appropriations, P/A/P level), FY 2018–2026; DBM-published execution aggregates (all available funds, agency level), FY 2011–2025. CHED has been an Other Executive Office (attached to the Office of the President) throughout the period. No FAR No. 1 (P/A/P-level execution) was available for CHED, so absorption is reported only at the agency aggregate, on a basis not directly comparable to the appropriations series; note that unspent free-tuition balances carry forward as continuing appropriations. Prepared as the long-form companion to the CHED budget briefing.
</p>
