---
title: "DPWH Budget Analysis"
subtitle: "Office of the Secretary — Long-Run Appropriations"
author: "AJ Montesa / People's Budget Coalition"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    md_extensions: "-smart"
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.width = 9, fig.height = 4.6, dpi = 200,
  out.width = "100%", fig.align = "center"
)
```

```{r libraries}
library(tidyverse)
library(readxl)
library(scales)
library(ggtext)
library(ggrepel)
library(knitr)
library(kableExtra)
```

```{r globals}
# --- File path -------------------------------------------------------------
# Path to the compiled data file (sourced from DBM publications).
data_path <- "../data/Compiled_-_DPWH.xlsx"

# --- Which series to deflate by --------------------------------------------
# Either "CPI" or "GDPDeflator" (both indexed to 2018 = 100).
deflator_choice <- "CPI"

# --- Colours ---------------------------------------------------------------
# Fixed mapping so NEP/GAA/Actuals are consistent across every plot.
item_levels <- c("NEP", "GAA", "Actual Obligations")
item_colors <- c("NEP" = "#ffa6b0", "GAA" = "#3994fc", "Actual Obligations" = "#bcbef2")

# --- DPWH programs (consistent labelling began with the 2018 budget) -------
dpwh_programs <- c("General Administration and Support", "Support to Operations",
                   "Asset Preservation Program", "Network Development Program",
                   "Bridge Program", "Flood Management Program",
                   "Local Program", "Convergence and Special Support Program")

# Eight-colour palette built around the house three (#bcbef2 / #ffa6b0 / #3994fc):
# cool tones for the engineering programs, warm tones for the discretionary ones.
program_colors <- c(
  "General Administration and Support"      = "#bcbef2",  # periwinkle (house)
  "Support to Operations"                   = "#c8a8e9",  # soft violet
  "Asset Preservation Program"              = "#3994fc",  # blue (house)
  "Network Development Program"             = "#6fcfd6",  # teal
  "Bridge Program"                          = "#8fd9a8",  # mint
  "Flood Management Program"                = "#ffe27a",  # warm yellow
  "Local Program"                           = "#ffcf8b",  # apricot
  "Convergence and Special Support Program" = "#ffa6b0"   # salmon (house)
)

# --- Philippine Congresses (for reference lines) ---------------------------
# Short ordinal labels so they don't crowd the 1988-2026 span.
congresses_ph2 <- tibble(
  Congress = c("8th", "9th", "10th", "11th", "12th", "13th", "14th",
               "15th", "16th", "17th", "18th", "19th", "20th"),
  Congress_Start = as.Date(c("1987-07-27", "1992-07-27", "1995-07-24", "1998-07-27",
                             "2001-07-23", "2004-07-26", "2007-07-23", "2010-07-26",
                             "2013-07-22", "2016-07-25", "2019-07-22", "2022-07-25",
                             "2025-07-28"))
)

# --- Presidential administrations (inauguration dates) ---------------------
administrations_ph <- tibble(
  President = c("C. Aquino", "Ramos", "Estrada", "Arroyo",
                "B. Aquino III", "Duterte", "Marcos Jr."),
  Start = as.Date(c("1986-02-25", "1992-06-30", "1998-06-30", "2001-01-20",
                    "2010-06-30", "2016-06-30", "2022-06-30"))
)

# Clamp the first era to the left edge of the data so its label stays on-canvas,
# and only draw boundary lines that fall inside the plotted range.
data_start      <- as.Date("1988-01-01")
admin_labels    <- administrations_ph %>% mutate(LabelX = pmax(Start, data_start))
admin_lines     <- administrations_ph %>% filter(Start >= data_start)
congress_labels <- congresses_ph2     %>% mutate(LabelX = pmax(Congress_Start, data_start))
congress_lines  <- congresses_ph2     %>% filter(Congress_Start >= data_start)

# --- Shared x-axis: tick + label ONLY the first budget year of each Congress
# (FY1988 = 8th, FY1993 = 9th, ... FY2020 = 18th, FY2023 = 19th, FY2026 = 20th)
congress_first_fy <- as.Date(c("1988-01-01", "1993-01-01", "1996-01-01", "1999-01-01",
                               "2002-01-01", "2005-01-01", "2008-01-01", "2011-01-01",
                               "2014-01-01", "2017-01-01", "2020-01-01", "2023-01-01",
                               "2026-01-01"))
```

# Overview

This first section profiles the long-run budget of the **DPWH Office of the
Secretary**: a (still incomplete) time series of the
**NEP**, **GAA**, and **Actual Obligations** running back to 1988. Three views
are shown — the **nominal** trend, the department's **share of the national
budget**, and the **real (inflation-adjusted)** trend in constant 2018 prices.

All figures throughout this document are drawn from the National Expenditure
Program (NEP), General Appropriations Act (GAA), and related budget documents
published by the Department of Budget and Management (DBM).

A few conventions:

- Amounts are reported in **thousand pesos**; everything below is
  rescaled to **PHP billions** (`/ 1e6`).
- The headline series is **New + Automatic Appropriations** (the full
  agency budget, including automatic items). This is also the numerator used
  for the share of the national budget.
- Points mark **years with observed data**; the connecting line is interpolated,
  so visible gaps flag where encoding is still pending.

```{r load-data}
# New + Automatic Appropriations (the headline DPWH OSec budget)
dpwh_newauto <- read_xlsx(data_path, sheet = "DPWH_OSec", range = "A2:D41") %>%
  rename(FY = 1, NEP = 2, GAA = 3, `Actual Obligations` = 4)

# Total national budget (denominator for DPWH's share)
total_budget <- read_xlsx(data_path, sheet = "DPWH_OSec", range = "K2:N41") %>%
  rename(FY = 1, NEP = 2, GAA = 3, `Actual Obligations` = 4)

# Price indices (2018 = 100)
price_index <- read_xlsx(data_path, sheet = "DPWH_OSec", range = "U2:W41") %>%
  rename(FY = 1, CPI = 2, GDPDeflator = 3)

# Tidy long table with nominal, real, and share columns
dpwh_long <- dpwh_newauto %>%
  pivot_longer(-FY, names_to = "Item", values_to = "Amount") %>%
  left_join(
    total_budget %>% pivot_longer(-FY, names_to = "Item", values_to = "TotalBudget"),
    by = c("FY", "Item")
  ) %>%
  left_join(price_index, by = "FY") %>%
  mutate(
    FY_date  = as_date(str_c(FY, "-01-01")),
    Item     = factor(Item, levels = item_levels),
    AmountB  = Amount / 1e6,                       # thousand pesos -> PHP billions
    Deflator = .data[[deflator_choice]],
    RealB    = AmountB / (Deflator / 100),         # constant 2018 prices
    Share    = Amount / TotalBudget
  )
```

# The Long View: DPWH Appropriations Since 1988

The four views below track the DPWH Office of the Secretary's budget across nearly
four decades — in nominal pesos, as a share of the national budget, in
inflation-adjusted terms, and in a single reference table — before the rest of
the document turns to the program, item, and district-level detail of recent years.

## Nominal Trend

```{r plot-nominal}
ggplot(data = dpwh_long %>% filter(!is.na(AmountB))) +
  geom_hline(yintercept = 0) +
  geom_vline(data = admin_lines, aes(xintercept = Start),
             linetype = "dashed", linewidth = 0.4, alpha = 0.45) +
  geom_vline(data = congress_lines, aes(xintercept = Congress_Start), alpha = 0.10) +
  geom_text(data = admin_labels, aes(x = LabelX, y = Inf, label = President),
            size = 2.4, fontface = "bold", hjust = -0.05, vjust = 1.2) +
  geom_text(data = congress_labels, aes(x = LabelX, y = Inf, label = Congress),
            size = 1.9, color = "gray55", hjust = -0.1, vjust = 2.9) +
  geom_path(aes(x = FY_date, y = AmountB, color = Item), linewidth = 1.1) +
  geom_point(aes(x = FY_date, y = AmountB, color = Item), size = 1.1) +
  scale_x_date(date_labels = "FY '%y", breaks = congress_first_fy, minor_breaks = NULL) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.05))) +
  scale_color_manual(values = item_colors) +
  coord_cartesian(ylim = c(0, NA)) +
  labs(x = NULL, y = "PHP billions", color = NULL,
       title = "DPWH Office of the Secretary: New + Automatic Appropriations",
       caption = "**Data Sources:** NEP and GAA documents, 1988-2026.<br>**Visualization:** AJ Montesa / People's Budget Coalition.") +
  theme_minimal() +
  theme(legend.position = c(0.15, 0.7),
        panel.grid.major.x = element_blank(),
        panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 13),
        plot.caption = element_markdown(size = 6),
        axis.text.x = element_text(size = 7))
```

In nominal terms the DPWH OSec budget climbs steeply from the mid-2010s. Two
deviations from the usual NEP-to-GAA pattern stand out and are worth raising in
hearings: the **2019 GAA fell below the NEP** (the year of the ₱95.4B line-item veto; a smaller ₱26.1B veto also trimmed the 2025 GAA),
and the **2026 GAA drops sharply below its NEP** — the reverse of the
usual upward realignment seen in most other years.

## Share of the National Budget

```{r plot-share}
ggplot(data = dpwh_long %>% filter(!is.na(Share))) +
  geom_hline(yintercept = 0) +
  geom_vline(data = admin_lines, aes(xintercept = Start),
             linetype = "dashed", linewidth = 0.4, alpha = 0.45) +
  geom_vline(data = congress_lines, aes(xintercept = Congress_Start), alpha = 0.10) +
  geom_text(data = admin_labels, aes(x = LabelX, y = Inf, label = President),
            size = 2.4, fontface = "bold", hjust = -0.05, vjust = 1.2) +
  geom_text(data = congress_labels, aes(x = LabelX, y = Inf, label = Congress),
            size = 1.9, color = "gray55", hjust = -0.1, vjust = 2.9) +
  geom_path(aes(x = FY_date, y = Share, color = Item), linewidth = 1.1) +
  geom_point(aes(x = FY_date, y = Share, color = Item), size = 1.1) +
  scale_x_date(date_labels = "FY '%y", breaks = congress_first_fy, minor_breaks = NULL) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                     expand = expansion(mult = c(0, 0.05))) +
  scale_color_manual(values = item_colors) +
  coord_cartesian(ylim = c(0, NA)) +
  labs(x = NULL, y = "% of Total Appropriations", color = NULL,
       title = "DPWH Office of the Secretary, as a Share of the National Budget",
       caption = "**Data Sources:** NEP and GAA documents, 1988-2026.<br>**Visualization:** AJ Montesa / People's Budget Coalition.") +
  theme_minimal() +
  theme(legend.position = c(0.13, 0.55),
        panel.grid.major.x = element_blank(),
        panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 13),
        plot.caption = element_markdown(size = 6),
        axis.text.x = element_text(size = 7))
```

The share view rescales the same money against the size of the whole budget.
The department's footprint roughly **doubles from ~6-8% (1990s-2000s) to a
mid-teens share** in the latest years — evidence that DPWH's expansion outpaced
overall budget growth, not merely tracked it. The 2026 NEP-to-GAA gap reappears
here as a one-year contraction in the GAA share.

## Real Trend (Inflation-Adjusted)

```{r plot-real}
ggplot(data = dpwh_long %>% filter(!is.na(RealB))) +
  geom_hline(yintercept = 0) +
  geom_vline(data = admin_lines, aes(xintercept = Start),
             linetype = "dashed", linewidth = 0.4, alpha = 0.45) +
  geom_vline(data = congress_lines, aes(xintercept = Congress_Start), alpha = 0.10) +
  geom_text(data = admin_labels, aes(x = LabelX, y = Inf, label = President),
            size = 2.4, fontface = "bold", hjust = -0.05, vjust = 1.2) +
  geom_text(data = congress_labels, aes(x = LabelX, y = Inf, label = Congress),
            size = 1.9, color = "gray55", hjust = -0.1, vjust = 2.9) +
  geom_path(aes(x = FY_date, y = RealB, color = Item), linewidth = 1.1) +
  geom_point(aes(x = FY_date, y = RealB, color = Item), size = 1.1) +
  scale_x_date(date_labels = "FY '%y", breaks = congress_first_fy, minor_breaks = NULL) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.05))) +
  scale_color_manual(values = item_colors) +
  coord_cartesian(ylim = c(0, NA)) +
  labs(x = NULL, y = "PHP billions (constant 2018 prices)", color = NULL,
       title = "DPWH Office of the Secretary, in Real Terms",
       subtitle = str_c("Deflated by the ",
                        if_else(deflator_choice == "CPI", "CPI", "GDP deflator"),
                        " (2018 = 100)"),
       caption = "**Data Sources:** NEP and GAA documents, 1988-2026; PSA price indices.<br>**Visualization:** AJ Montesa / People's Budget Coalition.") +
  theme_minimal() +
  theme(legend.position = c(0.15, 0.7),
        panel.grid.major.x = element_blank(),
        panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(size = 9),
        plot.caption = element_markdown(size = 6),
        axis.text.x = element_text(size = 7))
```

Holding prices constant at 2018 tempers the nominal story: much of the post-2018
"growth" is the price level catching up. In real terms the GAA budget is
**broadly flat between 2018 and 2023** before resuming its climb — a useful
counterweight to headline peso figures in the hearing room. Switch
`deflator_choice` to `"GDPDeflator"` in the setup chunk to see the
GDP-deflator version.

## Summary Table

```{r summary-table}
dpwh_table <- dpwh_long %>%
  mutate(code = recode(Item, "NEP" = "NEP", "GAA" = "GAA",
                       "Actual Obligations" = "AO")) %>%
  select(FY, code, AmountB, RealB, Share) %>%
  pivot_wider(names_from = code, values_from = c(AmountB, RealB, Share)) %>%
  transmute(
    `FY`                   = FY,
    `NEP`                  = AmountB_NEP,
    `GAA`                  = AmountB_GAA,
    `Actual Oblig.`        = AmountB_AO,
    `Realignment (GAA-NEP)`  = AmountB_GAA - AmountB_NEP,
    `DPWH Share (GAA)`     = Share_GAA,
    `GAA, real 2018`       = RealB_GAA
  ) %>%
  arrange(desc(FY))

# Pre-format so blanks read as "-" rather than NA
fmt_b <- function(x) { xn <- suppressWarnings(as.numeric(x)); ifelse(is.na(xn), "--", comma(xn, accuracy = 0.1)) }

dpwh_table %>%
  mutate(
    across(c(NEP, GAA, `Actual Oblig.`, `Realignment (GAA-NEP)`, `GAA, real 2018`), fmt_b),
    `DPWH Share (GAA)` = ifelse(is.na(suppressWarnings(as.numeric(`DPWH Share (GAA)`))), "--",
                                percent(`DPWH Share (GAA)`, accuracy = 0.1))
  ) %>%
  kbl(align = "lrrrrrr", escape = FALSE,
      caption = "DPWH Office of the Secretary, New + Automatic Appropriations (PHP billions unless noted)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 13)
```

# The DPWH Budget by Program

DPWH organises its budget into eight **programs**. Consistent program labelling
only begins with the **2018 budget**, so this section runs 2018-2026. These amounts are reported
in **pesos** and are rescaled to **PHP billions** (`/ 1e9`). All
figures are **New Appropriations** (the program totals reconcile to the
New-Appropriations column used in Section 1).

```{r load-byprogram}
byprogram <- read_xlsx(data_path, sheet = "NewAppro_ByProgram") %>%
  mutate(
    Vetoed  = replace_na(as.numeric(Vetoed), 0),
    Program = factor(Program, levels = dpwh_programs),
    NEP_B   = NEP / 1e9,
    GAA_B   = GAA / 1e9,                 # final (post-veto) GAA
    Veto_B  = Vetoed / 1e9,
    Realign = (GAA - NEP) / 1e9          # NEP -> enacted change (negative = cut)
  )

# Substantive programs (drops the two overhead lines) for the NEP-vs-GAA panels
substantive <- setdiff(dpwh_programs,
                       c("General Administration and Support", "Support to Operations"))
```

## Composition Over Time

```{r plot-composition-abs}
ggplot(byprogram, aes(x = FY, y = GAA_B, fill = Program)) +
  geom_col(color = "black", linewidth = 0.2, width = 0.82) +
  geom_vline(xintercept = 2022.5, linetype = "dashed", alpha = 0.45) +
  annotate("text", x = 2020,   y = Inf, label = "Duterte",
           vjust = 1.4, size = 2.6, fontface = "bold") +
  annotate("text", x = 2024.5, y = Inf, label = "Marcos Jr.",
           vjust = 1.4, size = 2.6, fontface = "bold") +
  scale_x_continuous(breaks = 2018:2026) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.07))) +
  scale_fill_manual(values = program_colors, labels = function(x) str_wrap(x, 26)) +
  labs(x = NULL, y = "PHP billions", fill = NULL,
       title = "DPWH Budget Composition by Program (GAA)",
       caption = "**Data Sources:** GAA documents, 2018-2026.<br>**Visualization:** AJ Montesa / People's Budget Coalition.") +
  guides(fill = guide_legend(ncol = 2)) +
  theme_minimal() +
  theme(legend.position = "bottom", legend.text = element_text(size = 6.5),
        legend.key.size = unit(0.35, "cm"),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold", size = 13),
        plot.caption = element_markdown(size = 6))
```

```{r plot-composition-share}
ggplot(byprogram, aes(x = FY, y = GAA_B, fill = Program)) +
  geom_col(color = "black", linewidth = 0.2, width = 0.82,
           position = position_fill()) +
  geom_vline(xintercept = 2022.5, linetype = "dashed", alpha = 0.45) +
  annotate("text", x = 2020,   y = Inf, label = "Duterte",
           vjust = 1.4, size = 2.6, fontface = "bold") +
  annotate("text", x = 2024.5, y = Inf, label = "Marcos Jr.",
           vjust = 1.4, size = 2.6, fontface = "bold") +
  scale_x_continuous(breaks = 2018:2026) +
  scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.07))) +
  scale_fill_manual(values = program_colors, labels = function(x) str_wrap(x, 26)) +
  labs(x = NULL, y = "Share of DPWH GAA", fill = NULL,
       title = "DPWH Budget Composition by Program (GAA, 100% stacked)",
       caption = "**Data Sources:** GAA documents, 2018-2026.<br>**Visualization:** AJ Montesa / People's Budget Coalition.") +
  guides(fill = guide_legend(ncol = 2)) +
  theme_minimal() +
  theme(legend.position = "bottom", legend.text = element_text(size = 6.5),
        legend.key.size = unit(0.35, "cm"),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold", size = 13),
        plot.caption = element_markdown(size = 6))
```

The composition shift is stark: the **Convergence and Special Support Program**
(apricot/salmon) swells under the current administration, while the **Network
Development** and **Flood Management** shares contract — most dramatically in the
**2026 GAA**, where Flood Management is cut to a sliver. The dashed line marks the
Duterte-to-Marcos transition (FY2023 is the first budget proposed by the current
administration).

```{r table-composition}
byprogram %>%
  transmute(FY, Program, GAA_B) %>%
  pivot_wider(names_from = FY, values_from = GAA_B) %>%
  arrange(factor(Program, levels = dpwh_programs)) %>%
  mutate(across(-Program, ~comma(.x, accuracy = 0.1))) %>%
  kbl(align = c("l", rep("r", 9)),
      caption = "GAA by Program, PHP billions") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 12)
```

## NEP vs GAA and Realignments

```{r plot-nepvgaa-faceted, fig.height=5.4}
nepvgaa <- byprogram %>%
  filter(Program %in% substantive, FY <= 2026) %>%
  mutate(Program = factor(Program, levels = substantive)) %>%
  select(FY, Program, NEP_B, GAA_B) %>%
  pivot_longer(c(NEP_B, GAA_B), names_to = "Item", values_to = "AmountB") %>%
  mutate(Item = factor(if_else(Item == "NEP_B", "NEP", "GAA"), levels = c("NEP", "GAA")))

nepvgaa_var <- byprogram %>%
  filter(Program %in% substantive, FY <= 2026) %>%
  mutate(Program = factor(Program, levels = substantive),
         ypos = pmax(NEP_B, GAA_B) * 1.04,
         lab  = if_else(Realign >= 0,
                        str_c("+", format(round(Realign, 1), nsmall = 1)),
                        format(round(Realign, 1), nsmall = 1)))

ggplot(nepvgaa) +
  geom_col(aes(x = FY, y = AmountB, fill = Item),
           position = position_dodge(width = 0.75), width = 0.66) +
  geom_text(data = nepvgaa_var, aes(x = FY, y = ypos, label = lab), size = 1.9) +
  facet_wrap(~Program, scales = "free_y", ncol = 3,
             labeller = label_wrap_gen(24)) +
  scale_x_continuous(breaks = c(2018, 2020, 2022, 2024, 2026)) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.13))) +
  scale_fill_manual(values = c("NEP" = "#ffa6b0", "GAA" = "#3994fc")) +
  labs(x = NULL, y = "PHP billions", fill = NULL,
       title = "DPWH by Program: NEP vs GAA, with Net Realignments (GAA - NEP)",
       caption = "**Data Sources:** NEP and GAA documents, 2018-2026. Labels show the net change from NEP to enacted GAA (PHP B).<br>**Visualization:** AJ Montesa / People's Budget Coalition.") +
  theme_bw() +
  theme(legend.position = "bottom", legend.text = element_text(size = 7),
        panel.grid.minor = element_blank(), panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold", size = 13),
        plot.caption = element_markdown(size = 6),
        strip.background = element_rect(fill = "white"),
        strip.text = element_text(size = 7),
        axis.text = element_text(size = 6.5))
```

The same realignments read most cleanly as a matrix — each cell is the net move
from the President's proposal (NEP) to the enacted law (GAA), with **blue =
increases** and **red = cuts**:

```{r plot-realign-heatmap, fig.height=4.2}
ggplot(byprogram %>% filter(FY <= 2026), aes(x = FY, y = fct_rev(Program), fill = Realign)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = round(Realign)), size = 2.1) +
  scale_x_continuous(breaks = 2018:2026, position = "top",
                     expand = expansion(mult = 0)) +
  scale_y_discrete(labels = function(x) str_wrap(x, 28)) +
  scale_fill_gradient2(low = "#ffa6b0", mid = "white", high = "#3994fc",
                       midpoint = 0, labels = scales::comma, name = "GAA - NEP\n(PHP B)") +
  labs(x = NULL, y = NULL,
       title = "Net Realignments from NEP to GAA, by Program",
       caption = "**Note:** 2019 and 2025 net out the P95.4B and P26.1B presidential vetoes respectively (see the veto section).<br>**Visualization:** AJ Montesa / People's Budget Coalition.") +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 13),
        plot.caption = element_markdown(size = 6),
        axis.text.y = element_text(size = 7),
        legend.key.width = unit(0.35, "cm"))
```

Two patterns stand out for the hearing room: the **Convergence and Special
Support Program** gains year after year (peaking at **+₱265B in 2025**),
while **Network Development** and **Bridge** programs are routinely trimmed from
their proposed levels. The **2026** column is exceptional — Flood Management is
realigned **down by ~₱246B**, the largest single-program cut in the series.

```{r table-realign}
byprogram %>%
  filter(FY <= 2026) %>%
  transmute(FY, Program, Realign) %>%
  pivot_wider(names_from = FY, values_from = Realign) %>%
  arrange(factor(Program, levels = dpwh_programs)) %>%
  mutate(across(-Program, ~comma(.x, accuracy = 0.1))) %>%
  kbl(align = c("l", rep("r", 9)),
      caption = "Net Realignment (GAA - NEP) by Program, PHP billions") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 12)
```

# Inside the Programs: Sub-Programs and Line Items

The program totals in Section 6 hide a great deal. Beneath each program sit
**sub-programs** and, below those, the individual **line items** — "Preventive
Maintenance – Primary Roads," "SIPAG – Access Roads... leading to Major/Strategic
Public Buildings," and dozens more — that are where budgeting actually happens.
This section works from a granular, multi-year encoding of those items (FY 2018
onward) to ask three questions the aggregates cannot answer: **where has emphasis
shifted between programs, what does Congress add to and cut at the most granular
level, and which line items are the biggest — and has that changed?**

For readability, line items that the budget splits only by road class — the
separate *Primary Roads*, *Secondary Roads* and *Tertiary Roads* versions of
Preventive Maintenance, Road Widening, and a few others — are combined here into a
single item each.

```{r granular-parse, include=FALSE}
parse_granular <- function(sheet, years) {
  raw <- read_xlsx(data_path, sheet = sheet, col_names = FALSE)
  hdr <- as.character(unlist(raw[1, ]))
  hdr[1:4] <- c("L0", "L1", "L2", "L3"); hdr[is.na(hdr)] <- paste0("x", which(is.na(hdr)))
  names(raw) <- make.unique(hdr)
  dat <- raw[-1, ]; dat$L3_orig <- dat$L3
  dat <- dat %>% mutate(across(c(L0, L1, L2, L3), as.character)) %>%
    fill(L0, .direction = "down") %>% fill(L1, .direction = "down") %>% fill(L2, .direction = "down")
  leaf <- dat %>% filter(!is.na(L3_orig))
  purrr::map_dfr(years, function(y) {
    nc <- paste0("NEP", y, "_Amount"); gc <- paste0("GAA_", y, "_Amount")
    leaf %>% transmute(Item = str_remove(L3, " - (Primary|Secondary|Tertiary) Roads$"),
      SubProg = L2, ProgramGroup = L1, Major = L0, FY = y,
      NEP = suppressWarnings(as.numeric(.data[[nc]])),
      GAA = if (gc %in% names(leaf)) suppressWarnings(as.numeric(.data[[gc]])) else NA_real_)
  }) %>%
    mutate(Program = if_else(ProgramGroup %in% c("Convergence and Special Support Program", "Local Program"),
                             ProgramGroup, SubProg),
           NEP_B = NEP / 1e9, GAA_B = GAA / 1e9, Add_B = GAA_B - NEP_B)
}
gran <- parse_granular("2018-2027", 2018:2027)
program_colors_short <- c("Asset Preservation" = "#3994fc", "Network Development" = "#6fcfd6",
                          "Bridge" = "#8fd9a8", "Flood Management" = "#ffe27a",
                          "Convergence & Special Support" = "#ffa6b0", "Local" = "#ffcf8b")
short_item <- function(x, wrap = 30) {
  x %>%
    # collapsed maintenance / rehab families
    str_replace("^Rehabilitation/ Reconstruction of Roads with Slips, Slope Collapse, and Landslide$",
                "Roads w/ Slips & Landslide (rehab)") %>%
    str_replace("^Rehabilitation/ Reconstruction/ Upgrading of Damaged Paved Roads$",
                "Rehab of Damaged Paved Roads") %>%
    str_replace("^Construction/ Upgrading/ Rehabilitation of Drainage along National Roads$",
                "Drainage along National Roads") %>%
    # flood management
    str_replace("Construction/Maintenance of Flood Mitigation Structures and Drainage Systems",
                "Flood Mitigation Structures & Drainage") %>%
    str_replace("Construction/Rehabilitation of Flood Mitigation Facilities within Major River Basins and Principal Rivers",
                "Flood Mitigation - Major River Basins") %>%
    # SIPAG / BIP families
    str_replace("SIPAG - Access Roads.*leading to Major/ Strategic Public Buildings/ Facilities", "SIPAG - Access Roads to Public Buildings") %>%
    str_replace("BIP - Access Roads.*leading to Major/ Strategic Public Buildings/ Facilities", "BIP - Access Roads to Public Buildings") %>%
    str_replace("(SIPAG|BIP) - Access Roads.*Indigenous People \\(IP\\) Communities", "\\1 - Access Roads to IP Communities") %>%
    str_replace("(SIPAG|BIP) - Interjurisdictional Roads.*", "\\1 - Interjurisdictional Roads") %>%
    str_replace("(SIPAG|BIP) - Coastal Roads/Causeway.*", "\\1 - Coastal Roads / Causeway") %>%
    str_replace("(SIPAG|BIP) - Flood Mitigation Structures protecting Public Infrastructures/Facilities", "\\1 - Flood Mitigation Structures") %>%
    str_replace("(SIPAG|BIP) - Multi-Purpose Buildings/ Facilities to support Social Services", "\\1 - Multi-Purpose Buildings") %>%
    str_replace("(SIPAG|BIP) - Ma[ij]or/Strategic Public Buildings.*", "\\1 - Strategic Public Buildings") %>%
    str_replace("(SIPAG|BIP) - Evacuation Centers.*", "\\1 - Evacuation / Health Facilities") %>%
    str_replace("(SIPAG|BIP) - Public Water Supply System", "\\1 - Public Water Supply") %>%
    str_replace("(SIPAG|BIP) - Local Ports and Boat Landings", "\\1 - Local Ports / Boat Landings") %>%
    str_replace("(SIPAG|BIP) - Climate-Resilient and Green Infrastructure Features", "\\1 - Climate-Resilient Infra") %>%
    # network / bridge
    str_replace("Construction of By-Pass and  Diversion Roads", "By-Pass & Diversion Roads") %>%
    str_replace("Construction of Missing Links/ New Roads", "Missing Links / New Roads") %>%
    str_replace("Construction of Flyovers/ Interchanges/ Underpasses/ Long Span Bridges", "Flyovers / Interchanges / Underpasses") %>%
    str_replace("Construction of New Bridges", "New Bridges") %>%
    str_replace("Replacement of Bridges \\(Temporary to Permanent\\)", "Replacement of Bridges (Temp to Perm)") %>%
    str_replace("Retrofitting/ Strengthening of Permanent Bridges", "Retrofitting of Permanent Bridges") %>%
    str_replace("Rehabilitation/ Major Repair of Permanent Bridges", "Major Repair of Permanent Bridges") %>%
    str_replace("Improvement of Capacity of Provincial Bridges along Provincial Roads", "Capacity of Provincial Bridges") %>%
    # access roads / misc
    str_replace("Construction/ Improvement of Access Roads leading to ", "Access Roads to ") %>%
    str_replace("Construction/ Rehabilitation of Water Supply/ Septage and Sewerage/ Rain Water Collectors", "Water Supply / Septage / Sewerage") %>%
    str_replace("Rehabilitation of Disaster-Related Infrastructure and Other Facilities", "Rehab of Disaster-Related Infra") %>%
    str_replace("Philippine Seismic Risk Reduction and Resilience Project.*", "Phil. Seismic Risk Reduction Project") %>%
    str_replace("Special Road Fund - Motor Vehicle User's Charge \\(MVUC\\) as per R.A. 11239", "Special Road Fund (MVUC)") %>%
    str_replace("^Multipurpose / Facilities$", "Multipurpose Facilities") %>%
    str_squish() %>%
    str_wrap(width = wrap)
}
```

## What's Inside the Major Programs

The program names in Section 6 are broad umbrellas. Below is what each of the six
operational programs actually contained in **FY2025** — the last full budget
before the flood-control scandal reshaped the FY2026 cycle — shown as its
constituent line items.

```{r plot-program-innards, fig.height=8.2}
prog6 <- c("Asset Preservation Program","Network Development Program","Bridge Program",
           "Flood Management Program","Convergence and Special Support Program","Local Program")
innards <- gran %>%
  filter(FY == 2025, Program %in% prog6) %>%
  group_by(Program, Item) %>% summarise(GAA_B = sum(GAA_B, na.rm = TRUE), .groups = "drop") %>%
  filter(GAA_B > 0) %>%
  group_by(Program) %>% slice_max(GAA_B, n = 8) %>% ungroup() %>%
  mutate(ProgLab = str_replace(Program, "Convergence and Special Support Program", "Convergence & Special Support") %>%
                   str_replace(" Program$",""),
         ProgLab = factor(ProgLab, levels = c("Asset Preservation","Network Development","Bridge",
                                              "Flood Management","Convergence & Special Support","Local")),
         key = reorder(str_c(short_item(Item), Program, sep = "___"), GAA_B))
ggplot(innards, aes(x = GAA_B, y = key, fill = ProgLab)) +
  geom_col(width = 0.72) +
  facet_wrap(~ProgLab, scales = "free", ncol = 2) +
  scale_y_discrete(labels = function(x) str_remove(x, "___.*")) +
  scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.10))) +
  scale_fill_manual(values = program_colors_short, guide = "none") +
  labs(x = "FY2025 enacted GAA (PHP billions)", y = NULL,
       title = "What each major program contains: largest line items, FY2025",
       subtitle = "Up to eight biggest items per program",
       caption = "**Data:** granular program encoding.<br>**Visualization:** AJ Montesa / People's Budget Coalition.") +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), panel.grid.major.y = element_blank(),
        strip.background = element_rect(fill = "white"), strip.text = element_text(face = "bold", size = 8.5),
        plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(size = 9), plot.caption = element_markdown(size = 6),
        axis.text.y = element_text(size = 6.3), axis.text.x = element_text(size = 6.5))
```

The contents follow the department's core mandates, with two important exceptions:

- **Asset Preservation** is road *upkeep*: preventive maintenance of primary,
  secondary and tertiary roads, plus rehabilitation of roads damaged by slips,
  landslides and disasters.
- **Network Development** is road *expansion*: road widening, by-pass and diversion
  roads, missing links and new roads, flyovers and interchanges.
- **Bridge** is exactly that — construction, replacement, widening and
  retrofitting of bridges.
- **Flood Management** is the government's headline flood line: construction and
  rehabilitation of flood-mitigation structures and drainage, including works
  along major river basins.
- **Convergence & Special Support** is the outlier — not a single purpose but a
  **basket** of two sub-programs, **SIPAG** and **BIP**, whose items are access
  roads to public buildings, multi-purpose buildings, *and their own flood-
  mitigation structures*. Much of what looks like "convergence" spending is thus
  district road-building, construction, and flood control under a different label.
- **Local Program** historically carried the same district works (local roads,
  multipurpose facilities, flood control and drainage) before FY2022, when most of
  it migrated into Convergence.

The key takeaway is that **flood-control money is spread across at least three
programs** — the Flood Management Program, the SIPAG/BIP flood-mitigation items
inside Convergence, and the Local Program's flood-control line — which is part of
why the total was so hard to see, and so easy to inflate, before the FY2026
reckoning.

## The Shift from the "Local Program" to "Convergence and Special Support"

The clearest structural change in this period is a **relabeling of district
infrastructure**. Through FY2021, locally-oriented projects ran through the
**Local Program**; the **Convergence and Special Support Program** was a modest
line. In **FY2022 the two swapped places** — Local collapsed and Convergence
absorbed its role, then grew far larger.

```{r plot-local-convergence, fig.height=4.2}
lc <- gran %>%
  filter(ProgramGroup %in% c("Convergence and Special Support Program", "Local Program")) %>%
  group_by(ProgramGroup, FY) %>%
  summarise(GAA = sum(GAA_B, na.rm = TRUE), NEP = sum(NEP_B, na.rm = TRUE), .groups = "drop") %>%
  mutate(Enacted = if_else(FY <= 2026, GAA, NA_real_),
         Group = recode(ProgramGroup,
                        "Convergence and Special Support Program" = "Convergence & Special Support",
                        "Local Program" = "Local Program"))
ggplot(lc, aes(x = FY, y = Enacted, colour = Group, group = Group)) +
  geom_line(linewidth = 1.1) + geom_point(size = 2.2) +
  annotate("rect", xmin = 2021.5, xmax = 2022.5, ymin = -Inf, ymax = Inf, alpha = 0.08, fill = "red") +
  annotate("text", x = 2022, y = max(lc$Enacted, na.rm = TRUE) * 0.55,
           label = "FY2022:\nroles swap", size = 2.9, fontface = "bold", colour = "grey30") +
  scale_x_continuous(breaks = 2018:2026) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.06))) +
  scale_colour_manual(values = c("Convergence & Special Support" = "#b8336a", "Local Program" = "#3994fc")) +
  labs(x = NULL, y = "Enacted GAA (PHP billions)", colour = NULL,
       title = "The Local Program gives way to Convergence & Special Support",
       subtitle = "Enacted new appropriations by program group, FY2018-2026",
       caption = "**Data:** granular program encoding, operations items only.<br>**Visualization:** AJ Montesa / People's Budget Coalition.") +
  theme_bw() +
  theme(legend.position = "top", panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(size = 9), plot.caption = element_markdown(size = 6))
```

Before the swap, the Local Program ran at **₱119-281B** while Convergence sat at
**₱30-62B**. In FY2022 the Local Program fell to **₱15B** and Convergence leapt to
**₱353B**; it peaked at **₱504B in FY2025** before the FY2026 flood-driven cut
brought it to **₱240B**. Convergence is not a single purpose but a **basket** —
it houses the **SIPAG** and **BIP** sub-programs, whose items range from access
roads and multi-purpose buildings to, notably, **flood-mitigation structures**.
So a large share of what looks like "convergence" or "special support" spending is,
underneath, the same district road-and-building-and-drainage work that the Local
Program used to carry — now concentrated in one fast-growing, discretionary
program.

## What Congress Adds To — and Cuts — at the Item Level

Reading Congress's changes (enacted minus proposed) at the **individual item
level**, over FY2018-2025 (the FY2026 cycle is treated separately below, as it was
reshaped by the flood-control scandal), reveals a consistent pattern: the legislature moves
money **out of large, national, strategic works and into district-scale,
insertable items**.

```{r plot-item-addcut, fig.height=5.6}
item_add <- gran %>%
  filter(FY <= 2025) %>%
  group_by(Item, Program) %>%
  summarise(Add = sum(Add_B, na.rm = TRUE), .groups = "drop")
addcut <- bind_rows(
  item_add %>% slice_max(Add, n = 10),
  item_add %>% slice_min(Add, n = 6)
) %>%
  mutate(Item = short_item(Item),
         Dir = if_else(Add >= 0, "Added by Congress", "Cut by Congress"),
         Item = fct_reorder(Item, Add))
ggplot(addcut, aes(x = Add, y = Item, fill = Dir)) +
  geom_col(width = 0.72) +
  geom_text(aes(label = sprintf("%+.0f", Add), hjust = if_else(addcut$Add >= 0, -0.12, 1.12)),
            size = 2.8, colour = "grey20") +
  scale_fill_manual(values = c("Added by Congress" = "#1a9850", "Cut by Congress" = "#d73027")) +
  scale_x_continuous(labels = function(v) paste0(v, "B"), expand = expansion(mult = c(0.14, 0.14))) +
  labs(x = "Cumulative enacted minus proposed, FY2018-2025 (PHP billions)", y = NULL, fill = NULL,
       title = "The line items Congress builds up --- and the ones it cuts",
       subtitle = "Top 10 additions and 6 largest cuts, summed over FY2018-2025",
       caption = "**Data:** granular program encoding.<br>**Visualization:** AJ Montesa / People's Budget Coalition.") +
  theme_bw() +
  theme(legend.position = "top", panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(size = 9), plot.caption = element_markdown(size = 6),
        axis.text.y = element_text(size = 7.5))
```

The additions are dominated by **SIPAG and BIP** items — *Access Roads to Public
Buildings* (**+₱207B** and **+₱105B**), *Multi-Purpose Buildings* (**+₱176B**),
and their *Flood-Mitigation Structures* variants — together with **Local Roads
(+₱182B)** and **Multipurpose Facilities (+₱104B)**. The cuts fall on **large
national works**: *By-Pass & Diversion Roads* (**-₱94B**), *Flood Mitigation in
Major River Basins* (**-₱71B**), *New Bridges* (**-₱49B**), and *National Roads*
(**-₱43B**). In other words, the legislature consistently trims the department's
big strategic-network and major-river projects and redirects the money into
smaller, district-level, more easily-sited items — the granular signature of the
Central-Office-to-districts pattern documented in the operating-unit analysis that
follows.

```{r table-top-item-adds}
item_add %>%
  slice_max(Add, n = 15) %>%
  transmute(`Line item` = short_item(Item), Program = str_remove(Program, " Program$"),
            `Cumulative addition (PHP B)` = round(Add, 1)) %>%
  arrange(desc(`Cumulative addition (PHP B)`)) %>%
  kbl(align = c("l", "l", "r"),
      caption = "Line items Congress most enlarged, cumulative FY2018-2025 (enacted minus proposed)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE, font_size = 11) %>%
  row_spec(0, bold = TRUE)
```

## The FY2026 Cycle: A Scandal-Driven Revision

FY2026 was unlike any other year in this dataset, and it is why the analysis above
stops at FY2025. Midway through 2025 a **flood-control corruption scandal** broke
open — ghost projects, kickbacks, and a wave of identically-priced "multi-purpose
building" and flood-control insertions. The fallout ran straight through the budget.

The original FY2026 NEP, prepared under then-Secretary Manuel Bonoan and submitted
in August 2025, proposed the department's usual large flood-control program. After
the scandal, a new Secretary — **Vivencio "Vince" Dizon** — took over in
September and submitted an **errata to the NEP**, removing locally-funded
flood-control projects on the President's directive. Congress then cut further.
The result is a rare three-step collapse: **original NEP → revised NEP → enacted
GAA**, all *downward*, the reverse of the department's usual pattern.

```{r plot-2026-revision, fig.height=4.6}
# Recompute the three FY2026 stages (incl. the revised-NEP errata column) from the sheet
rev_raw <- read_xlsx(data_path, sheet = "2018-2027", col_names = FALSE)
hdrx <- as.character(unlist(rev_raw[1,])); hdrx[1:4] <- c("L0","L1","L2","L3"); hdrx[is.na(hdrx)] <- paste0("z",which(is.na(hdrx)))
names(rev_raw) <- make.unique(hdrx)
rev_raw <- rev_raw[-1,]; rev_raw$L3o <- rev_raw$L3
rev_raw <- rev_raw %>% mutate(across(c(L0,L1,L2,L3),as.character)) %>%
  fill(L0,.direction="down") %>% fill(L1,.direction="down") %>% fill(L2,.direction="down") %>% filter(!is.na(L3o)) %>%
  mutate(Program = if_else(L1 %in% c("Convergence and Special Support Program","Local Program"), L1, L2),
         onep = suppressWarnings(as.numeric(NEP2026_Amount))/1e9,
         rnep = suppressWarnings(as.numeric(NEPrev2026_Amount))/1e9,
         gaa  = suppressWarnings(as.numeric(GAA_2026_Amount))/1e9)
rev26 <- rev_raw %>% group_by(Program) %>%
  summarise(`Original NEP`=sum(onep,na.rm=TRUE), `Revised NEP`=sum(rnep,na.rm=TRUE), `Enacted GAA`=sum(gaa,na.rm=TRUE), .groups="drop") %>%
  filter(`Original NEP`+`Revised NEP`+`Enacted GAA` > 1)
rev26_long <- rev26 %>% pivot_longer(-Program, names_to="Stage", values_to="B") %>%
  mutate(Stage=factor(Stage, levels=c("Original NEP","Revised NEP","Enacted GAA")),
         ProgLab=str_replace(Program,"Convergence and Special Support Program","Convergence & Spec. Support") %>% str_replace(" Program$",""),
         ProgLab=fct_reorder(ProgLab, B, .fun=max, .desc=TRUE))
ggplot(rev26_long, aes(x=B, y=fct_rev(ProgLab), fill=Stage)) +
  geom_col(position=position_dodge(width=0.78), width=0.72) +
  scale_x_continuous(labels=scales::comma, expand=expansion(mult=c(0,0.06))) +
  scale_fill_manual(values=c("Original NEP"="#f4a582","Revised NEP"="#b8336a","Enacted GAA"="#3994fc")) +
  labs(x="PHP billions", y=NULL, fill=NULL,
       title="FY2026: Original NEP -> Revised NEP (errata) -> Enacted GAA, by Program",
       subtitle="The errata gutted the Flood Management Program; Congress then cut Network Development further",
       caption="**Data:** granular encoding (NEP, revised-NEP errata, and GAA columns). Operations items only.<br>**Visualization:** AJ Montesa / People's Budget Coalition.") +
  theme_bw() +
  theme(legend.position="top", panel.grid.minor=element_blank(), panel.grid.major.y=element_blank(),
        plot.title=element_text(face="bold", size=12.5), plot.subtitle=element_text(size=8.5),
        plot.caption=element_markdown(size=6), axis.text.y=element_text(size=8))
```

Two moves stand out. The **Flood Management Program was gutted in the errata itself**
— from about ₱251B in the original NEP to roughly ₱16B in the revised version, a
cut of some ₱235B before Congress even voted — as the flood-mitigation items at
the heart of the scandal were pulled. But the revision was not purely subtractive:
the **Convergence & Special Support Program actually rose** in the revised NEP
(from ~₱168B to ~₱225B) and again in the enacted GAA (~₱240B), even as everything
around it shrank. Congress then trained its knife on **Network Development**,
cutting it from ~₱149B in the revised NEP to ~₱97B enacted.

So the headline "DPWH budget cut by roughly a third" masks a reshuffle: the
scandal removed the flood-control program that had been the vehicle for insertions,
but the discretionary **Convergence** basket — which, as shown above, contains its
*own* flood-mitigation and district-building items — emerged from the FY2026
process *larger*, not smaller. Whether that represents genuine reform or the same
spending under a cleaner-sounding label is exactly the kind of question the
granular view is meant to surface.

## The Biggest Line Items, and How They've Shifted

Ranking items by **enacted** size shows how the composition of DPWH spending has
turned over. In FY2018 the largest items were classic **road and flood works**; by
FY2025 the **SIPAG and BIP** items — which barely existed earlier — sit at the
very top, alongside the two big flood-mitigation lines.

```{r plot-biggest-items-shift, fig.height=4.8}
big_cmp <- gran %>%
  filter(FY %in% c(2018, 2025)) %>%
  group_by(FY, Item) %>% summarise(GAA_B = sum(GAA_B, na.rm = TRUE), .groups = "drop") %>%
  group_by(FY) %>% slice_max(GAA_B, n = 10) %>% ungroup() %>%
  mutate(ItemS = short_item(Item),
         FYlab = factor(paste0("FY", FY), levels = c("FY2018", "FY2025")),
         key = reorder(str_c(ItemS, FY, sep = "___"), GAA_B))
ggplot(big_cmp, aes(x = GAA_B, y = key, fill = FYlab)) +
  geom_col(width = 0.72) +
  geom_text(aes(label = round(GAA_B)), hjust = -0.15, size = 2.7, colour = "grey20") +
  facet_wrap(~FYlab, scales = "free", ncol = 2) +
  scale_y_discrete(labels = function(x) str_remove(x, "___.*")) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  scale_fill_manual(values = c("FY2018" = "#9ecae1", "FY2025" = "#08519c"), guide = "none") +
  labs(x = "Enacted GAA (PHP billions)", y = NULL,
       title = "The 10 biggest line items: FY2018 vs FY2025",
       subtitle = "SIPAG and BIP items rise from nothing to the top of the budget",
       caption = "**Data:** granular program encoding.<br>**Visualization:** AJ Montesa / People's Budget Coalition.") +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), panel.grid.major.y = element_blank(),
        strip.background = element_rect(fill = "white"), strip.text = element_text(face = "bold"),
        plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(size = 9), plot.caption = element_markdown(size = 6),
        axis.text.y = element_text(size = 7))
```

The turnover is striking. In FY2018 the biggest items were **Flood Mitigation
Structures (₱97B)**, **Road Widening (₱64B)**, **By-Pass & Diversion Roads
(₱57B)**, **Local Roads (₱51B)** and **Missing Links / New Roads (₱36B)** — a
portfolio of national road and flood works. By FY2025, the **SIPAG Access Roads to
Public Buildings (₱137B)** and **BIP Multi-Purpose Buildings (₱96B)** items — which
did not exist under those names before FY2022 — rank among the very largest,
alongside the two big flood-mitigation lines (which Congress simultaneously *cut*
from the proposal) and, once the road classes are combined, **Preventive
Maintenance (₱104B)** — the quiet, recurring cost of keeping the existing network
usable. The composition of the
DPWH budget, in short, has tilted from building and maintaining the national
network toward a fast-expanding, district-oriented "convergence" portfolio.



# New Appropriations by Operating Unit

The DPWH disburses its budget across some 200+ **operating units** — the Central
Office, regional offices, and district engineering offices (DEOs). The data
run from the **FY2014 budget** through **FY2026**, spanning the 16th through 20th
Congresses — the Aquino, Duterte, and Marcos administrations, and **seven House
Speakers**. The series is now complete across all thirteen fiscal years; the GAA
encoded for **2019** reflects the budget *after* the President's ₱95.4-billion
line-item veto (isolated below). That span lets us trace
how district money has moved
between proposal and enactment across administrations. Amounts are in **pesos**,
rescaled to **PHP billions**.

```{r load-byoperunit}
congress_colors <- c("16th (Aquino)"  = "#c9ccd1",
                     "17th (Duterte)" = "#ffcf8b",
                     "18th (Duterte)" = "#ffa6b0",
                     "19th (Marcos)"  = "#3994fc",
                     "20th (Marcos)"  = "#7fd0d6")

# Per-fiscal-year palette: one coherent hue family per Congress, shaded from the
# oldest year (light) to the newest (dark). Grey is reserved for the vetoed amounts.
fy_colors <- c(
  "2014" = "#fbe7a4", "2015" = "#f3cd4e", "2016" = "#d4a417",   # 16th (Aquino)  -- yellow
  "2017" = "#f6b0a8", "2018" = "#e36253", "2019" = "#c22d22",   # 17th (Duterte) -- red
  "2020" = "#ddd0ee", "2021" = "#b89ad9", "2022" = "#8a5cc0",   # 18th (Duterte) -- purple
  "2023" = "#b9d4f6", "2024" = "#6fa6e8", "2025" = "#2f6fd0",   # 19th (Marcos)  -- blue
  "2026" = "#5cc0c4")                                           # 20th (Marcos)  -- teal

byoperunit <- read_xlsx(data_path, sheet = "NewAppro_ByOperUnit") %>%
  mutate(FY = as.integer(FY),
         # The FY2027 NEP-by-office rows are encoded with each office's UACS code
         # in the FY column (>= 100000); relabel them to FY 2027.
         FY = if_else(FY >= 100000, 2027L, FY)) %>%
  # 2019's encoded GAA is the post-veto (final authorized) budget; the P95.4B
  # line-item veto is isolated in the Central-vs-District veto chart below.
  mutate(NEP     = replace_na(NEP, 0),
         # FY2027 is proposed only (no GAA yet); keep its GAA missing, not zero,
         # so it never reads as a realignment down to zero.
         GAA     = if_else(FY == 2027L, NA_real_, replace_na(GAA, 0)),
         GAA_B   = GAA / 1e9,
         NEP_B   = NEP / 1e9,
         Realign = (GAA - NEP) / 1e9,
         Congress = case_when(FY %in% 2014:2016 ~ "16th (Aquino)",
                              FY %in% 2017:2019 ~ "17th (Duterte)",
                              FY %in% 2020:2022 ~ "18th (Duterte)",
                              FY %in% 2023:2025 ~ "19th (Marcos)",
                              FY %in% 2026:2027 ~ "20th (Marcos)"),
         Admin    = case_when(FY %in% 2014:2016 ~ "Aquino",
                              FY %in% 2017:2022 ~ "Duterte",
                              FY %in% 2023:2027 ~ "Marcos"),
         # House Speaker presiding when each FY budget was deliberated and passed
         Speaker  = case_when(FY %in% 2014:2016 ~ "Belmonte",
                              FY %in% 2017:2018 ~ "Alvarez",
                              FY == 2019        ~ "Arroyo",
                              FY == 2020        ~ "Cayetano",
                              FY %in% 2021:2022 ~ "Velasco",
                              FY %in% 2023:2025 ~ "Romualdez",
                              FY %in% 2026:2027 ~ "Dy"))

# District engineering offices. `deo` covers enacted years (FY2017-2026) and
# feeds every realignment/veto view unchanged; `deo_all` additionally carries the
# FY2027 proposal for the levels section below.
deo_all <- byoperunit %>% filter(str_detect(Office, "D.E. Office"), FY >= 2017)
deo <- deo_all %>% filter(FY <= 2026)
n_years_deo <- dplyr::n_distinct(deo$FY)
```

Because the realignments that move money toward district engineering offices are made on the
House floor, the relevant political actor is often the **Speaker of the day** as
much as the sitting President. **Seven** Speakers presided over the budgets in this
window; their terms and the fiscal-year budgets each steered to passage are below.

```{r speaker-table}
tibble::tribble(
  ~Speaker,                  ~`Term began`,   ~Congress, ~`FY budgets`,  ~Administration,
  "Feliciano Belmonte Jr.",  "22 Jul 2013",   "16th",    "FY2014-2016",  "Aquino",
  "Pantaleon Alvarez",       "25 Jul 2016",   "17th",    "FY2017-2018",  "Duterte",
  "Gloria Macapagal Arroyo", "23 Jul 2018",   "17th",    "FY2019*",      "Duterte",
  "Alan Peter Cayetano",     "22 Jul 2019",   "18th",    "FY2020",       "Duterte",
  "Lord Allan Velasco",      "12 Oct 2020",   "18th",    "FY2021-2022",  "Duterte",
  "Martin Romualdez",        "25 Jul 2022",   "19th",    "FY2023-2025",  "Marcos",
  'Faustino "Bojie" Dy III', "17 Sep 2025",   "20th",    "FY2026",       "Marcos"
) %>%
  kbl(align = "lllll",
      caption = "House Speakers presiding over the DPWH budgets analysed here. *FY2019's encoded GAA is post-veto; the ₱95.4B line-item veto that year fell mostly on the Central Office (see the veto chart below).") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 12)
```

## Most of the Money Sits in the Central Office

```{r plot-central, fig.height=4}
central_split <- byoperunit %>%
  filter(FY <= 2026) %>%
  mutate(Unit = if_else(Office == "Central Office", "Central Office", "All Other Offices")) %>%
  group_by(FY, Unit) %>%
  summarize(GAA_B = sum(GAA_B), .groups = "drop") %>%
  mutate(Unit = factor(Unit, levels = c("All Other Offices", "Central Office")))

central_lab <- central_split %>%
  group_by(FY) %>%
  summarize(tot = sum(GAA_B),
            central = sum(GAA_B[Unit == "Central Office"]), .groups = "drop") %>%
  mutate(lab = percent(central / tot, accuracy = 1))

ggplot(central_split, aes(x = FY, y = GAA_B, fill = Unit)) +
  geom_col(color = "black", linewidth = 0.2, width = 0.78) +
  geom_text(data = central_lab, aes(x = FY, y = tot, label = lab),
            inherit.aes = FALSE, vjust = -0.5, size = 2.8) +
  scale_x_continuous(breaks = c(2014, 2017, 2020, 2023, 2026)) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.09))) +
  scale_fill_manual(values = c("All Other Offices" = "#bcbef2", "Central Office" = "#3994fc")) +
  labs(x = NULL, y = "GAA, PHP billions", fill = NULL,
       title = "Most of the DPWH Budget Is Lodged in the Central Office",
       subtitle = "Labels show the Central Office's share of the GAA",
       caption = "**Data Sources:** GAA documents, 2014-2026.<br>**Visualization:** AJ Montesa / People's Budget Coalition.") +
  theme_minimal() +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank(), panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(size = 9),
        plot.caption = element_markdown(size = 6))
```

Across the series the Central Office holds most of the *enacted* budget —
typically **60-75%**, peaking at **75% in 2025** — with 2015-2016
(both under 50%) and the unusual 2026 GAA (62%) the main exceptions. A single operating unit carrying most
of the department's budget is a reason for *closer* scrutiny of centrally-managed
projects, not a reason to assume they are already well accounted for. The enacted
shares also understate how concentrated the executive's *proposal* is, and how
much Congress rearranges it on the way to enactment — the subject of the section
that follows.

```{r table-central}
central_split %>%
  pivot_wider(names_from = Unit, values_from = GAA_B) %>%
  mutate(Total = `Central Office` + `All Other Offices`,
         `Central share` = percent(`Central Office` / Total, accuracy = 0.1)) %>%
  transmute(FY,
            `Central Office (B)`    = comma(`Central Office`, accuracy = 0.1),
            `All Other Offices (B)` = comma(`All Other Offices`, accuracy = 0.1),
            `Total (B)`             = comma(Total, accuracy = 0.1),
            `Central share`) %>%
  kbl(align = "lrrrr",
      caption = "DPWH GAA: Central Office vs all other operating units, PHP billions") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 12)
```

## From Proposal to Law: Central Office to the Districts

The executive's proposal concentrates the budget in the Central Office even more
than the enacted law does. In pesos the redistribution is large and one-directional:
Congress channels tens — and, in recent years, hundreds — of billions toward
district engineering offices between the proposal and the enacted law, in **every
year on record**, while the Central Office is cut in most (first chart). In share
terms the pattern is cleaner still: in every year (2014-2026) the Central
Office's share of new
appropriations *falls* from NEP to GAA while the districts' share *rises* — a
swing reaching nearly **30 percentage points** in 2021 and 2023, when the proposal
parked almost the entire budget centrally (second chart). Congress, in other words,
systematically moves money toward named district engineering offices; this is the redistribution
that the office-by-office realignment analysis below traces in detail.

```{r plot-central-shift, fig.height=4.2}
byoperunit %>%
  filter(FY <= 2026) %>%
  mutate(Unit = case_when(Office == "Central Office"        ~ "Central Office",
                          str_detect(Office, "D.E. Office") ~ "District Engineering Offices",
                          TRUE                              ~ "Other")) %>%
  filter(Unit != "Other") %>%
  group_by(FY, Unit) %>%
  summarize(Shift_B = sum(GAA_B) - sum(NEP_B), .groups = "drop") %>%
  ggplot(aes(x = factor(FY), y = Shift_B, fill = Unit)) +
  geom_col(position = position_dodge(width = 0.72), width = 0.66,
           color = "black", linewidth = 0.2) +
  geom_hline(yintercept = 0, linewidth = 0.4) +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_manual(values = c("Central Office" = "#3994fc", "District Engineering Offices" = "#bcbef2")) +
  labs(x = NULL, y = "Change in budget, NEP -> GAA\n(PHP billions)", fill = NULL,
       title = "Congress Redirects the Budget Toward the District Engineering Offices",
       subtitle = "NEP-to-GAA change in each block's budget: district engineering offices gain every year; the Central Office is cut in most",
       caption = "**Data Sources:** NEP and GAA documents, 2014-2026.<br>**Visualization:** AJ Montesa / People's Budget Coalition.") +
  theme_minimal() +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank(), panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(size = 9),
        plot.caption = element_markdown(size = 6))
```

```{r plot-central-levels, fig.height=4}
byoperunit %>%
  filter(FY <= 2026) %>%
  mutate(Unit = case_when(Office == "Central Office"        ~ "Central Office",
                          str_detect(Office, "D.E. Office") ~ "District Engineering Offices",
                          TRUE                              ~ "Other")) %>%
  group_by(FY, Unit) %>%
  summarize(NEP_B = sum(NEP_B), GAA_B = sum(GAA_B), .groups = "drop") %>%
  group_by(FY) %>%
  mutate(`Proposed (NEP)` = NEP_B / sum(NEP_B),
         `Enacted (GAA)`  = GAA_B / sum(GAA_B)) %>%
  ungroup() %>%
  filter(Unit != "Other") %>%
  pivot_longer(c(`Proposed (NEP)`, `Enacted (GAA)`), names_to = "Stage", values_to = "Share") %>%
  mutate(Stage = factor(Stage, levels = c("Proposed (NEP)", "Enacted (GAA)"))) %>%
  ggplot(aes(x = factor(FY), y = Share, fill = Stage)) +
  geom_col(position = position_dodge(width = 0.72), width = 0.66,
           color = "black", linewidth = 0.2) +
  facet_wrap(~Unit) +
  scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.06))) +
  scale_fill_manual(values = c("Proposed (NEP)" = "#ffa6b0", "Enacted (GAA)" = "#3994fc")) +
  labs(x = NULL, y = "Share of new appropriations", fill = NULL,
       title = "Share of New Appropriations: Proposed vs Enacted",
       subtitle = "The Central Office is proposed a larger share than it is enacted; for district engineering offices it is the reverse",
       caption = "**Data Sources:** NEP and GAA documents, 2014-2026.<br>**Visualization:** AJ Montesa / People's Budget Coalition.") +
  theme_minimal() +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank(), panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(size = 9),
        plot.caption = element_markdown(size = 6),
        axis.text.x = element_text(size = 7),
        strip.text = element_text(face = "bold", size = 9))
```

Two of those Central cuts come with an asterisk, because they are not really
Congress's doing. The operating-unit GAA encoded for **2019** is the budget *after*
the President's **₱95.4-billion line-item veto** — the largest in recent memory
— and that veto landed overwhelmingly on the Central Office (**₱74B** of the ₱95B
struck, versus ₱21B across the districts). That is why 2019 shows such a deep
Central cut above. Congress's own pre-veto budget was routine: it trimmed the
Central Office by roughly **₱26B** and added about **₱30B** to the districts, the
usual direction of travel. The chart below splits what Congress actually passed into
the part that survived the veto (blue) and the part the President struck out (grey);
the diamonds mark the President's original proposal (NEP). Because the veto hit the
Central Office hardest, it paradoxically *raised* the district share of the final
budget (from 38% pre-veto to 42% enacted).

```{r plot-veto-operunit, fig.height=4}
veto_unit <- byoperunit %>%
  filter(FY == 2019) %>%
  mutate(Unit = case_when(Office == "Central Office"        ~ "Central Office",
                          str_detect(Office, "D.E. Office") ~ "District Engineering Offices",
                          TRUE                              ~ "Other")) %>%
  group_by(Unit) %>%
  summarize(NEP_B = sum(NEP_B), Enacted_B = sum(GAA_B), .groups = "drop") %>%
  left_join(
    read_xlsx(data_path, sheet = "2019_Veto") %>%
      mutate(Unit = case_when(Office == "Central Office"        ~ "Central Office",
                              str_detect(Office, "D.E. Office") ~ "District Engineering Offices",
                              TRUE                              ~ "Other")) %>%
      group_by(Unit) %>%
      summarize(Vetoed_B = sum(GAA_Veto) / 1e9, .groups = "drop"),
    by = "Unit"
  ) %>%
  filter(Unit != "Other") %>%
  mutate(Vetoed_B = replace_na(Vetoed_B, 0))

veto_unit %>%
  select(Unit, `Enacted (survived)` = Enacted_B, Vetoed = Vetoed_B) %>%
  pivot_longer(c(`Enacted (survived)`, Vetoed), names_to = "Stage", values_to = "Amount_B") %>%
  mutate(Stage = factor(Stage, levels = c("Vetoed", "Enacted (survived)"))) %>%
  ggplot(aes(x = Unit, y = Amount_B, fill = Stage)) +
  geom_col(width = 0.6, color = "black", linewidth = 0.2) +
  geom_point(data = veto_unit, aes(x = Unit, y = NEP_B),
             inherit.aes = FALSE, shape = 18, size = 4, color = "#ffa6b0") +
  geom_text(data = veto_unit,
            aes(x = Unit, y = Enacted_B + Vetoed_B / 2,
                label = paste0("vetoed\n", round(Vetoed_B), "B")),
            inherit.aes = FALSE, size = 2.6, color = "grey25", lineheight = 0.9) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.1))) +
  scale_fill_manual(values = c("Enacted (survived)" = "#3994fc", "Vetoed" = "grey70")) +
  labs(x = NULL, y = "PHP billions", fill = NULL,
       title = "The 2019 Line-Item Veto, by Operating Unit",
       subtitle = "Bars show what Congress passed; grey is the part the President struck. Diamonds mark the proposal (NEP).",
       caption = "**Data Sources:** NEP, GAA, and FY2019 veto message.<br>**Visualization:** AJ Montesa / People's Budget Coalition.") +
  theme_minimal() +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank(), panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(size = 9),
        plot.caption = element_markdown(size = 6),
        axis.text.x = element_text(size = 9))
```

## The Largest District Engineering Offices in Each Congress

Which districts command the biggest budgets is not fixed — it shifts from one
Congress to the next as offices are created, priorities move, and realignment
patterns change. The charts and tables below rank the **top 20 district
engineering offices in each Congress**, showing both what the Executive
**proposed (NEP)** and what was finally **enacted (GAA)**, summed over that
Congress's budgets. Reading down the Congresses gives a sense of which parts of
the country the DPWH budget has most favoured over time.

```{r deo-top20-bycongress, fig.height=12}
# Top 20 DEOs by enacted GAA within each Congress (enacted years only, FY<=2026),
# with the matching proposed (NEP) amount shown alongside. Ranked independently
# per facet using the Office___Congress key trick.
deo_cong <- deo %>%
  group_by(Congress, Office) %>%
  summarise(NEP = sum(NEP_B, na.rm = TRUE),
            GAA = sum(GAA_B, na.rm = TRUE), .groups = "drop")

top20_cong <- deo_cong %>%
  group_by(Congress) %>% slice_max(GAA, n = 20) %>% ungroup()

cong_label <- c("17th (Duterte)" = "17th Congress (FY2017-2019)",
                "18th (Duterte)" = "18th Congress (FY2020-2022)",
                "19th (Marcos)"  = "19th Congress (FY2023-2025)",
                "20th (Marcos)"  = "20th Congress (FY2026)")

top20_cong %>%
  pivot_longer(c(NEP, GAA), names_to = "Item", values_to = "AmountB") %>%
  mutate(Item      = factor(Item, levels = c("NEP", "GAA")),
         CongLabel = factor(recode(Congress, !!!cong_label), levels = unname(cong_label)),
         OfficeKey = reorder(str_c(Office, Congress, sep = "___"), AmountB * (Item == "GAA"))) %>%
  ggplot(aes(x = AmountB, y = OfficeKey, fill = Item)) +
  geom_col(position = position_dodge(width = 0.72), width = 0.68) +
  facet_wrap(~CongLabel, ncol = 2, scales = "free") +
  scale_y_discrete(labels = function(x) str_remove(x, "___.*")) +
  scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.06))) +
  scale_fill_manual(values = c("NEP" = "#ffa6b0", "GAA" = "#3994fc")) +
  labs(x = "PHP billions (summed over the Congress)", y = NULL, fill = NULL,
       title = "Top 20 District Engineering Offices in Each Congress: Proposed vs Enacted",
       subtitle = "Ranked by enacted GAA within each Congress; pink = proposed (NEP), blue = enacted (GAA)",
       caption = "**Data Sources:** NEP and GAA documents. The 20th Congress covers FY2026 only (FY2027 is proposed and shown in the levels section below).<br>**Visualization:** AJ Montesa / People's Budget Coalition.") +
  theme_minimal() +
  theme(legend.position = "top",
        panel.grid.major.y = element_blank(), panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(size = 9),
        plot.caption = element_markdown(size = 6),
        strip.text = element_text(size = 9, face = "bold"),
        axis.text.y = element_text(size = 6.5),
        axis.text.x = element_text(size = 6.5))
```

A few patterns stand out. The offices that top the list migrate over time —
**Abra** and other Cordillera and northern-Luzon districts dominate the Duterte
Congresses, while **Bulacan**, **Nueva Ecija**, and Metro Manila offices rise to
the top under Marcos. And in almost every case the **enacted (blue) bar towers
over the proposed (pink) one**: the largest districts are large not because the
Executive proposed them so, but because Congress built them up on enactment —
the realignment story, seen from the districts' side.

### Top 20 tables, by Congress

The same rankings in full, with the exact proposed and enacted amounts and the net
addition Congress made on enactment. Each table is the **top 20 offices by enacted
GAA** for that Congress.

```{r deo-top20-tables, results="asis"}
make_top20 <- function(cong, label) {
  tb <- deo_cong %>%
    filter(Congress == cong) %>%
    slice_max(GAA, n = 20) %>%
    arrange(desc(GAA)) %>%
    transmute(Rank = row_number(),
              `District Engineering Office` = str_remove(Office, " D\\.E\\. Office$"),
              `NEP (PHP B)` = round(NEP, 2),
              `GAA (PHP B)` = round(GAA, 2),
              `Net add (PHP B)` = round(GAA - NEP, 2))
  cat(sprintf("\n\n#### %s\n\n", label))
  print(
    kbl(tb, align = c("r", "l", "r", "r", "r"),
        caption = sprintf("Top 20 District Engineering Offices by Enacted GAA --- %s", label)) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = FALSE, font_size = 12)
  )
}
invisible(Map(make_top20, names(cong_label), unname(cong_label)))
```


## Top District Engineering Offices, by Size and by Realignment

Setting the Central Office aside, two questions matter for the hearings: which
district engineering offices command the **largest budgets**, and how much of each budget was
in the President's proposal versus added on enactment. The first two charts break
the largest offices into what was **proposed** (grey) and what Congress **added** on
enactment — shown as a single total (purple) in the first chart, then with the
additions split between the **Duterte** (red) and **Marcos** (blue) administrations in
the second. The third chart ranks offices by the realignments they received within
each Congress — now reaching back to the 16th (the three Aquino-era budgets) —
coloured by the fiscal year that delivered each addition, with the amounts the
President later vetoed shown in grey — the FY2019 vetoes on the 17th-Congress bars
(medium grey) and the smaller FY2025 vetoes on the 19th (dark grey).

```{r plot-deo-proposed-added, fig.height=5.4}
pva <- deo %>%
  group_by(Office) %>%
  summarize(`In proposal (NEP)` = sum(NEP_B),
            `Added by Congress` = sum(Realign),
            GAA_B = sum(GAA_B), .groups = "drop") %>%
  slice_max(GAA_B, n = 15)

pva %>%
  pivot_longer(c(`In proposal (NEP)`, `Added by Congress`),
               names_to = "Component", values_to = "AmountB") %>%
  mutate(Component = factor(Component, levels = c("In proposal (NEP)", "Added by Congress")),
         Office    = factor(Office, levels = arrange(pva, GAA_B)$Office)) %>%
  ggplot(aes(x = AmountB, y = Office, fill = Component)) +
  geom_col(color = "black", linewidth = 0.2) +
  scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.05))) +
  scale_fill_manual(values = c("In proposal (NEP)" = "grey75", "Added by Congress" = "#9a82c9")) +
  labs(x = "PHP billions, cumulative FY2017-2026", y = NULL, fill = NULL,
       title = "District Engineering Offices with the Largest Enacted Budgets",
       subtitle = "Grey = executive proposal (NEP); purple = added by Congress on enactment",
       caption = "**Data Sources:** NEP and GAA documents, 2017-2026.<br>**Visualization:** AJ Montesa / People's Budget Coalition.") +
  theme_minimal() +
  theme(legend.position = c(0.8, 0.3),
        panel.grid.major.y = element_blank(), panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(size = 9),
        plot.caption = element_markdown(size = 6),
        axis.text.y = element_text(size = 7))
```

```{r plot-deo-proposed-added-congress, fig.height=5.4}
pva_rank <- deo %>% filter(FY != 2026) %>%
  group_by(Office) %>% summarize(GAA_B = sum(GAA_B), .groups = "drop") %>%
  slice_max(GAA_B, n = 15)

pva_seg <- bind_rows(
  deo %>% filter(FY != 2026) %>% group_by(Office) %>%
    summarize(AmountB = sum(NEP_B), .groups = "drop") %>%
    mutate(Component = "In proposal (NEP)"),
  deo %>% filter(FY != 2026) %>% group_by(Office, Admin) %>%
    summarize(AmountB = sum(Realign), .groups = "drop") %>%
    transmute(Office, Component = str_c("Added under ", Admin), AmountB)
)

pva_seg %>%
  semi_join(pva_rank, by = "Office") %>%
  mutate(Office    = factor(Office, levels = arrange(pva_rank, GAA_B)$Office),
         Component = factor(Component, levels = c("In proposal (NEP)",
                                                  "Added under Duterte",
                                                  "Added under Marcos"))) %>%
  ggplot(aes(x = AmountB, y = Office, fill = Component)) +
  geom_col(color = "black", linewidth = 0.2) +
  scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.05))) +
  scale_fill_manual(values = c("In proposal (NEP)"  = "grey75",
                               "Added under Duterte" = "#e15759",
                               "Added under Marcos"  = "#3994fc")) +
  labs(x = "PHP billions, cumulative FY2017-2025", y = NULL, fill = NULL,
       title = "Proposed Budget and What Each Administration Added",
       subtitle = "FY2026 omitted: the one year Congress cut, rather than added to, district budgets",
       caption = "**Data Sources:** NEP and GAA documents, 2017-2025.<br>**Visualization:** AJ Montesa / People's Budget Coalition.") +
  theme_minimal() +
  theme(legend.position = c(0.74, 0.26),
        panel.grid.major.y = element_blank(), panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(size = 9),
        plot.caption = element_markdown(size = 6),
        axis.text.y = element_text(size = 7))
```

```{r plot-deo-insertions-fy, fig.height=12}
ins_fy <- byoperunit %>%
  filter(str_detect(Office, "D.E. Office"), FY <= 2025) %>%
  group_by(Office, Congress, FY) %>%
  summarize(Insertion = sum(Realign), .groups = "drop") %>%
  group_by(Office, Congress) %>%
  mutate(Insertion_total = sum(Insertion)) %>%
  ungroup()

# Top 12 offices within each Congress, ranked independently per facet
top_per_congress <- ins_fy %>%
  distinct(Office, Congress, Insertion_total) %>%
  group_by(Congress) %>%
  slice_max(Insertion_total, n = 12) %>%
  ungroup()

plotted <- ins_fy %>%
  semi_join(top_per_congress, by = c("Office", "Congress")) %>%
  mutate(Seg = as.character(FY)) %>%
  select(Office, Congress, Insertion, Insertion_total, Seg)

# FY2019 amounts Congress passed but the President vetoed, overlaid (grey) on the
# 17th-Congress bars so Congress's intent and the enacted total are both visible
veto_rows <- read_xlsx(data_path, sheet = "2019_Veto") %>%
  filter(str_detect(Office, "D.E. Office")) %>%
  group_by(Office) %>%
  summarize(Insertion = sum(GAA_Veto) / 1e9, .groups = "drop") %>%
  mutate(Congress = "17th (Duterte)", Seg = "Vetoed (2019)") %>%
  semi_join(filter(top_per_congress, Congress == "17th (Duterte)"),
            by = c("Office", "Congress")) %>%
  left_join(distinct(plotted, Office, Congress, Insertion_total),
            by = c("Office", "Congress"))

# FY2025 DEO-level vetoes, overlaid (a darker grey) on the 19th-Congress bars, the
# same way the 2019 amounts are shown on the 17th. The 2025 veto was overwhelmingly
# a Central Office event, so only small district slivers appear here.
veto_rows_25 <- read_xlsx(data_path, sheet = "2025_Veto") %>%
  filter(str_detect(Office, "D.E. Office")) %>%
  group_by(Office) %>%
  summarize(Insertion = sum(GAA_Veto) / 1e9, .groups = "drop") %>%
  mutate(Congress = "19th (Marcos)", Seg = "Vetoed (2025)") %>%
  semi_join(filter(top_per_congress, Congress == "19th (Marcos)"),
            by = c("Office", "Congress")) %>%
  left_join(distinct(plotted, Office, Congress, Insertion_total),
            by = c("Office", "Congress"))

seg_levels <- c(as.character(c(2014, 2015, 2016, 2017, 2018, 2019, 2020:2025)), "Vetoed (2019)", "Vetoed (2025)")
seg_colors <- c(fy_colors, "Vetoed (2019)" = "grey55", "Vetoed (2025)" = "grey30")

bind_rows(plotted, veto_rows, veto_rows_25) %>%
  mutate(
    CongLabel = recode(Congress,
                       "16th (Aquino)"  = "16th Congress (2014-2016)",
                       "17th (Duterte)" = "17th Congress (2017-2019)",
                       "18th (Duterte)" = "18th Congress (2020-2022)",
                       "19th (Marcos)"  = "19th Congress (2023-2025)"),
    # unique key per facet so each panel orders its own offices by total insertion
    OfficeKey = reorder(str_c(Office, Congress, sep = "___"), Insertion_total),
    Seg = factor(Seg, levels = seg_levels)
  ) %>%
  ggplot(aes(x = Insertion, y = OfficeKey, fill = Seg)) +
  geom_col(color = "black", linewidth = 0.2, position = position_stack(reverse = TRUE)) +
  facet_wrap(~CongLabel, ncol = 1, scales = "free_y") +
  scale_y_discrete(labels = function(x) str_remove(x, "___.*")) +
  scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0.04, 0.05))) +
  scale_fill_manual(values = seg_colors, drop = FALSE) +
  guides(fill = guide_legend(ncol = 1)) +
  labs(x = "PHP billions (realignment over the Congress)", y = NULL, fill = NULL,
       title = "District Engineering Offices Receiving the Largest Realignments, by Congress",
       subtitle = "Bars stack oldest year (left) to newest; grey = amounts Congress passed that the President vetoed (medium grey = 2019, dark grey = 2025)",
       caption = "**Data Sources:** NEP and GAA documents, 2014-2025.<br>**Visualization:** AJ Montesa / People's Budget Coalition.") +
  theme_minimal() +
  theme(legend.position = "right",
        panel.grid.major.y = element_blank(), panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(size = 9),
        plot.caption = element_markdown(size = 6),
        strip.text = element_text(size = 9, face = "bold"),
        axis.text.y = element_text(size = 7))
```

The realignment matrix below splits each of the most-favoured offices' realignments
between the **Duterte** and **Marcos** administrations — making clear, for instance,
that **Abra**'s windfalls came overwhelmingly under Duterte (2017-2022), while others
(e.g. **Bulacan 1st**, **Tarlac**) lean Marcos.

```{r plot-deo-heatmap, fig.height=4.6}
top15 <- deo %>% group_by(Office) %>%
  summarize(tot = sum(Realign), .groups = "drop") %>% slice_max(tot, n = 15) %>% pull(Office)

deo %>%
  filter(Office %in% top15) %>%
  mutate(Office = factor(Office, levels = rev(top15))) %>%
  ggplot(aes(x = FY, y = Office, fill = Realign)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = format(round(Realign, 1), nsmall = 1)), size = 2) +
  scale_x_continuous(breaks = 2017:2026, position = "top", expand = expansion(mult = 0)) +
  scale_fill_gradient2(low = "#ffa6b0", mid = "white", high = "#3994fc",
                       midpoint = 0, labels = scales::comma, name = "GAA - NEP\n(PHP B)") +
  labs(x = NULL, y = NULL,
       title = "Net Realignments by District Engineering Office and Year",
       caption = "**Data Sources:** NEP and GAA documents, 2017-2026.<br>**Visualization:** AJ Montesa / People's Budget Coalition.") +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", size = 13),
        plot.caption = element_markdown(size = 6),
        axis.text.y = element_text(size = 7),
        legend.key.width = unit(0.35, "cm"))
```

```{r table-deo-realign}
deo %>%
  group_by(Office, Admin) %>%
  summarize(Realign = sum(Realign), .groups = "drop") %>%
  pivot_wider(names_from = Admin, values_from = Realign, values_fill = 0) %>%
  mutate(Total = Duterte + Marcos) %>%
  slice_max(Total, n = 20) %>%
  arrange(desc(Total)) %>%
  transmute(Office,
            `Duterte (B)` = comma(Duterte, accuracy = 0.1),
            `Marcos (B)`  = comma(Marcos, accuracy = 0.1),
            `Total (B)`   = comma(Total, accuracy = 0.1)) %>%
  kbl(align = "lrrr",
      caption = "Top 20 D.E. Offices by total NEP-to-GAA realignment, FY2017-2026 (PHP billions)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 12)
```

## Which District Engineering Offices Merit a Closer Look?

Rather than single out any one region, this section applies a transparent screen
to *every* district engineering office and lets the data nominate which ones deserve a deeper
dive. Each office is scored on four independent signals over FY2017-2026 (the
Duterte and Marcos administrations):

- **Size** — it is among the ten largest recipients of realignments;
- **Intensity** — realignments make up at least 60% of the office's enacted budget;
- **Persistence** — it received a realignment in at least eight of the ten years;
- **Tilt** — the office's realignment totals under the two administrations differ by at least ₱4B.

Every signal builds on the same yearly *realignment* — the gap between an office's
enacted and proposed budget, **GAA − NEP** (in ₱B). The tilt signal sums those yearly
realignments within each administration and takes the difference between the two totals:

**admin tilt = (total GAA − NEP across FY2017-2022) − (total GAA − NEP across FY2023-2026),**

i.e. the office's total Duterte-era realignment minus its total Marcos-era realignment. A
positive tilt means the office gained more under Duterte and is shown **red** in the chart
below; a negative tilt means it gained more under Marcos and is shown **blue**; values near
zero are grey.

Offices that trip several signals at once are the natural candidates for scrutiny.

```{r deo-screen-data}
screen_base <- deo %>%
  group_by(Office) %>%
  summarize(NEP_B         = sum(NEP_B),
            GAA_B         = sum(GAA_B),
            Insertion     = sum(Realign),
            InsShare      = sum(Realign) / sum(GAA_B),
            YearsInserted = sum(Realign > 0.5),
            .groups = "drop") %>%
  left_join(
    deo %>%
      group_by(Office, Admin) %>% summarize(r = sum(Realign), .groups = "drop") %>%
      pivot_wider(names_from = Admin, values_from = r, values_fill = 0) %>%
      transmute(Office, Duterte, Marcos, Tilt = Duterte - Marcos),
    by = "Office"
  ) %>%
  mutate(across(c(Duterte, Marcos, Tilt), ~replace_na(.x, 0)))

# Newly-created district engineering offices: nothing proposed across the window (NEP = 0),
# so the realignment share is 100% by construction. Set aside and note in prose.
screen_nonep <- screen_base %>% filter(NEP_B == 0) %>% arrange(desc(GAA_B))

deo_screen <- screen_base %>%
  filter(NEP_B > 0) %>%
  mutate(
    flag_big        = min_rank(desc(Insertion)) <= 10,
    flag_intense    = InsShare >= 0.60,
    flag_persistent = YearsInserted >= n_years_deo - 2,
    flag_tilted     = abs(Tilt) >= 4,
    nflags = flag_big + flag_intense + flag_persistent + flag_tilted,
    Signals = str_c(
      if_else(flag_big,                    "large; ", ""),
      if_else(flag_intense,                "realignment-heavy; ", ""),
      if_else(flag_persistent,             "every year; ", ""),
      if_else(flag_tilted & Tilt > 0,      "Duterte-leaning; ", ""),
      if_else(flag_tilted & Tilt < 0,      "Marcos-leaning; ", "")
    ),
    Signals = str_remove(Signals, "; $")
  )

# District engineering offices whose realignment share falls below the plotted window (net trims)
screen_offscale <- deo_screen %>%
  filter(InsShare < 0) %>%
  arrange(InsShare)

offscale_note <- if (nrow(screen_offscale) == 0) {
  "none do"
} else {
  str_c(screen_offscale$Office, " (", round(screen_offscale$InsShare * 100), "%)",
        collapse = "; ")
}

nonep_note <- if (nrow(screen_nonep) == 0) {
  "none"
} else {
  str_c(screen_nonep$Office, collapse = "; ")
}
```

```{r plot-deo-screen, fig.height=5}
ggplot(deo_screen, aes(x = Insertion, y = InsShare)) +
  geom_point(aes(size = YearsInserted, fill = Tilt), shape = 21, color = "black", alpha = 0.9) +
  geom_text_repel(data = deo_screen %>% filter(nflags >= 2 | Insertion >= 9),
                  aes(label = Office), size = 2.3, max.overlaps = 30, segment.alpha = 0.4) +
  scale_fill_gradient2(low = "#3994fc", mid = "grey88", high = "#ff5d6c", midpoint = 0,
                       labels = scales::comma,
                       name = "Admin tilt, PHP B\n(red = Duterte,\nblue = Marcos)") +
  scale_size(range = c(1, 6), name = "Years with a\nrealignment") +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::percent) +
  coord_cartesian(ylim = c(0, NA)) +
  labs(x = "Total realignment, FY2017-2026 (PHP billions)",
       y = "Realignments as a share of the office's enacted budget",
       title = "Screening District Engineering Offices: Size and Intensity of Realignments",
       subtitle = "Top-right = both large and realignment-heavy; colour shows which administration leaned in",
       caption = "**Data Sources:** NEP and GAA documents, 2017-2026.<br>**Visualization:** AJ Montesa / People's Budget Coalition.") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(size = 9),
        plot.caption = element_markdown(size = 6),
        legend.title = element_text(size = 7), legend.text = element_text(size = 7))
```

The vertical axis starts at 0% and extends only as far as the data require. Two sets of offices fall outside it. First, `r nrow(screen_nonep)` newly-created district engineering offices had **no proposed budget** (NEP = 0) anywhere in the FY2017-2026 window, so their realignment share is 100% by construction — money Congress provided where the executive proposed nothing — and they are excluded from the scatter: `r nonep_note`. Second, `r nrow(screen_offscale)` very small offices fall *below* 0%, where the cumulative *enacted* budget came in under the proposal (a net trim rather than the padding that dominates the chart): `r offscale_note`. Because these offices are tiny, a modest peso cut reads as a large negative share, so those values are denominator artefacts more than meaningful signals.

The screen returns a consistent shortlist — none of it about a single region.
**Batangas 4th D.E. Office** is the largest single recipient of realignments —
**₱24.3B**, or **69%** of its enacted budget, padded in **all ten years** and
favoured fairly evenly by both administrations (₱14.5B Duterte, ₱9.7B Marcos), a
standing fixture regardless of who holds power. **Sorsogon** is similarly
relentless (₱16.0B, 66%, every year) but tilts Marcos-ward. **Bulacan 1st**
(₱20.7B) is the second-largest recipient and leans Marcos (₱12.7B vs ₱7.9B). At
the politically lopsided extremes sit **Abra** — the clearest Duterte-era office
(₱13.2B vs ₱3.1B) — and **Metro Manila 3rd**, its Marcos-era mirror (₱10.7B vs
₱3.8B). These offices, not any one province, are where the strongest patterns lie
and where project-level records would be worth pulling.

```{r table-deo-screen}
deo_screen %>%
  arrange(desc(nflags), desc(Insertion)) %>%
  slice_head(n = 15) %>%
  transmute(
    Office,
    `Realignment (B)`   = comma(Insertion, accuracy = 0.1),
    `Realignment share` = percent(InsShare, accuracy = 1),
    `Years realigned` = str_c(YearsInserted, "/", n_years_deo),
    `Duterte (B)`     = comma(Duterte, accuracy = 0.1),
    `Marcos (B)`      = comma(Marcos, accuracy = 0.1),
    Flags             = nflags,
    Signals
  ) %>%
  kbl(align = "lrrcrrcl",
      caption = "District engineering offices flagged for a closer look, sorted by number of signals then size") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 12) %>%
  column_spec(8, width = "13em")
```

# The FY 2027 Proposal

The FY 2027 National Expenditure Program is the live budget document now before
Congress — the one advocates can still shape. The proposal's figures already run
through the tables and charts in the sections above; this section draws them
together in one place: the proposal's overall size, its distribution **by program**
and **by district engineering office**, and how it compares to the two most useful
benchmarks — what the Executive proposed for FY 2026 (the previous **NEP**) and
what Congress actually enacted (the previous **GAA**).

## FY 2027 in Context: Comparison with FY 2026

At **₱642.6B**, the FY 2027 proposal sits between the two FY 2026 figures. It is
about **₱113B _above_** the ₱529.6B that was finally enacted for 2026 — a budget
slashed by the roughly ₱350B flood-control cut — but about **₱237B _below_** the
₱880B the Executive had originally proposed for 2026. In other words, the
Department is proposing to spend more than the shrunken 2026 law allows, but well
short of its own ambitions a year earlier.

```{r plot-2627-compare, fig.height=5.2}
cmp <- byprogram %>%
  filter(FY %in% c(2026, 2027)) %>%
  transmute(Program,
            `NEP 2026` = if_else(FY == 2026, NEP_B, NA_real_),
            `GAA 2026` = if_else(FY == 2026, GAA_B, NA_real_),
            `NEP 2027` = if_else(FY == 2027, NEP_B, NA_real_)) %>%
  group_by(Program) %>%
  summarise(across(everything(), ~ sum(.x, na.rm = TRUE)), .groups = "drop") %>%
  pivot_longer(-Program, names_to = "Series", values_to = "AmountB") %>%
  mutate(Series = factor(Series, levels = c("NEP 2026", "GAA 2026", "NEP 2027")),
         Program = factor(Program, levels = dpwh_programs))

ggplot(cmp, aes(x = AmountB, y = fct_rev(Program), fill = Series)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7) +
  scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.06))) +
  scale_y_discrete(labels = function(x) str_wrap(x, 24)) +
  scale_fill_manual(values = c("NEP 2026" = "#ffd0d5", "GAA 2026" = "#3994fc", "NEP 2027" = "#c0392b")) +
  labs(x = "PHP billions", y = NULL, fill = NULL,
       title = "FY 2027 Proposal vs FY 2026, by Program",
       subtitle = "Light pink = FY2026 proposed; blue = FY2026 enacted; red = FY2027 proposed",
       caption = "**Data Sources:** NEP and GAA documents.<br>**Visualization:** AJ Montesa / People's Budget Coalition.") +
  theme_bw() +
  theme(legend.position = "top", panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(size = 9), plot.caption = element_markdown(size = 6),
        axis.text.y = element_text(size = 8))
```

Read program by program, the FY 2027 proposal looks far more like the FY 2026
*proposal* than the FY 2026 *enacted* budget — the Executive is, in effect,
re-proposing much of what Congress rearranged on the way to the 2026 GAA:

- **Flood Management**, gutted to just **₱4.8B** in the 2026 GAA, is restored to
  **₱103.5B** in the FY 2027 proposal (though still below its ₱250.8B FY2026 NEP).
- **Network Development**, which Congress cut to **₱96.8B** on enactment, returns
  to **₱176.6B** — close to its original FY2026 proposal.
- The **Convergence and Special Support Program** — which Congress *inflated* to
  **₱240.5B** on enactment — is proposed back down at **₱157.3B**, roughly its
  FY2026 proposed level.

The full comparison, with the proposed change measured against the enacted 2026
budget (the current baseline):

```{r table-2627-compare}
byprogram %>%
  filter(FY %in% c(2026, 2027)) %>%
  group_by(Program) %>%
  summarise(`NEP 2026` = sum(NEP_B[FY == 2026], na.rm = TRUE),
            `GAA 2026` = sum(GAA_B[FY == 2026], na.rm = TRUE),
            `NEP 2027` = sum(NEP_B[FY == 2027], na.rm = TRUE), .groups = "drop") %>%
  mutate(`Chg. vs 2026 GAA` = `NEP 2027` - `GAA 2026`,
         Program = factor(Program, levels = dpwh_programs)) %>%
  arrange(desc(`NEP 2027`)) %>%
  mutate(across(where(is.numeric), ~ round(.x, 1))) %>%
  kbl(align = c("l", "r", "r", "r", "r"),
      caption = "FY 2027 Proposed New Appropriations vs FY 2026, by Program (PHP billions)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 12) %>%
  row_spec(0, bold = TRUE)
```


## By Program: Proposed and Enacted Levels

Zooming in on the programs, it helps to see the **overall levels** the Executive
proposes and the legislature enacts — the raw distribution of pesos across
programs, and where the **FY 2027 proposal** places them relative to the recent
past.

At the department level, the FY 2027 NEP proposes **₱642.6B** — a marked step
down from the ₱880B proposed for 2026 and the ₱1,087.7B *enacted* for 2025 (the
2025 figure is the post-veto total; see the veto section). The proposed-versus-enacted
totals over the full window make the swings plain:

```{r plot-levels-total, fig.height=3.8}
totlevels <- byprogram %>%
  group_by(FY) %>%
  summarise(NEP = sum(NEP_B, na.rm = TRUE),
            GAA = if (all(is.na(GAA_B))) NA_real_ else sum(GAA_B, na.rm = TRUE),
            .groups = "drop") %>%
  pivot_longer(c(NEP, GAA), names_to = "Item", values_to = "AmountB") %>%
  filter(!is.na(AmountB)) %>%
  mutate(Item = factor(Item, levels = c("NEP", "GAA")))

ggplot(totlevels, aes(x = FY, y = AmountB, colour = Item, group = Item)) +
  geom_line(linewidth = 0.9) + geom_point(size = 1.8) +
  geom_text(aes(label = round(AmountB)), vjust = -0.8, size = 2.4, show.legend = FALSE) +
  scale_x_continuous(breaks = c(seq(2018, 2026, 2), 2027)) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0.05, 0.14))) +
  scale_colour_manual(values = c("NEP" = "#ffa6b0", "GAA" = "#3994fc")) +
  labs(x = NULL, y = "PHP billions", colour = NULL,
       title = "DPWH Total New Appropriations: Proposed (NEP) vs Enacted (GAA)",
       caption = "**Note:** FY2027 shows the NEP only (not yet enacted). FY2025 GAA is the post-veto total.<br>**Visualization:** AJ Montesa / People's Budget Coalition.") +
  theme_bw() +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold", size = 13),
        plot.caption = element_markdown(size = 6))
```

The enacted budget ran *above* the proposal every year from 2020 to 2025 — a
legislature adding to the President's plan — until FY2026, when the enacted
total fell far *below* the proposal (the ₱246B flood-control cut). The FY2027
proposal resets the level near ₱643B.

**How is the FY 2027 proposal distributed across programs?** Two programs account
for more than half of it: Network Development (₱176.6B, 27%) and the
Convergence and Special Support Program (₱157.3B, 25%) — the latter
historically the *discretionary* program that Congress inflated through
realignments, now proposed at a high level in the Executive's own document.

```{r plot-levels-byprogram, fig.height=5.0}
lvl <- byprogram %>%
  select(FY, Program, NEP_B, GAA_B) %>%
  pivot_longer(c(NEP_B, GAA_B), names_to = "Item", values_to = "AmountB") %>%
  mutate(Item = factor(if_else(Item == "NEP_B", "NEP (proposed)", "GAA (enacted)"),
                       levels = c("NEP (proposed)", "GAA (enacted)"))) %>%
  filter(!is.na(AmountB))

ggplot(lvl, aes(x = FY, y = AmountB, fill = Program)) +
  geom_col(width = 0.72) +
  facet_wrap(~Item, ncol = 1) +
  scale_x_continuous(breaks = c(seq(2018, 2026, 2), 2027)) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.04))) +
  scale_fill_manual(values = program_colors) +
  labs(x = NULL, y = "PHP billions", fill = NULL,
       title = "DPWH Budget Levels by Program: Proposed (NEP, to 2027) and Enacted (GAA, to 2026)",
       caption = "**Note:** FY2027 GAA is omitted (not yet enacted). Each bar is the full department budget, split by program.<br>**Visualization:** AJ Montesa / People's Budget Coalition.") +
  theme_bw() +
  theme(legend.position = "bottom", legend.text = element_text(size = 6.5),
        panel.grid.minor = element_blank(), panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold", size = 12),
        plot.caption = element_markdown(size = 6),
        strip.background = element_rect(fill = "white"),
        strip.text = element_text(size = 9, face = "bold")) +
  guides(fill = guide_legend(nrow = 2))
```

Reading the two panels together is instructive. In the **enacted** budgets (lower
panel), the Convergence and Special Support Program balloons after 2022 — the
realignment story documented in Section 6. But the **proposed** budgets (upper
panel) show the Executive itself steadily raising Convergence, so that by FY2027
it is proposed at ₱157B before Congress touches it. The proposed FY2027
distribution, in full:

```{r table-levels-2027}
byprogram %>%
  filter(FY == 2027) %>%
  transmute(Program,
            `NEP 2027 (PHP B)` = round(NEP_B, 1),
            `Share of proposal` = scales::percent(NEP_B / sum(NEP_B), accuracy = 0.1)) %>%
  arrange(desc(`NEP 2027 (PHP B)`)) %>%
  kbl(align = c("l", "r", "r"),
      caption = "FY 2027 Proposed New Appropriations by Program (NEP)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 12)
```


## Inside the Proposal: The Biggest Line Items

Applying the same granular lens to the **FY 2027 proposal** shows what the Executive
is prioritising *within* the programs. The largest proposed items are a telling mix
of the national works Congress usually trims and the district-oriented items it
usually adds.

```{r plot-2027-items, fig.height=5.4}
items27 <- gran %>%
  filter(FY == 2027, NEP_B > 0) %>%
  group_by(Item, Program) %>% summarise(NEP_B = sum(NEP_B, na.rm = TRUE), .groups = "drop") %>%
  slice_max(NEP_B, n = 15) %>%
  mutate(ProgLab = str_replace(Program, "Convergence and Special Support Program", "Convergence & Special Support") %>%
                   str_replace(" Program$",""),
         ItemS = short_item(Item), ItemS = fct_reorder(ItemS, NEP_B))
ggplot(items27, aes(x = NEP_B, y = ItemS, fill = ProgLab)) +
  geom_col(width = 0.74) +
  geom_text(aes(label = round(NEP_B)), hjust = -0.15, size = 2.7, colour = "grey20") +
  scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.10))) +
  scale_fill_manual(values = program_colors_short, name = NULL) +
  labs(x = "FY 2027 proposed NEP (PHP billions)", y = NULL,
       title = "The 15 biggest line items in the FY 2027 proposal",
       subtitle = "Coloured by program; a mix of national road works and Convergence district items",
       caption = "**Data:** granular program encoding.<br>**Visualization:** AJ Montesa / People's Budget Coalition.") +
  theme_bw() +
  theme(legend.position = "bottom", legend.text = element_text(size = 6.5),
        panel.grid.minor = element_blank(), panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(size = 9), plot.caption = element_markdown(size = 6),
        axis.text.y = element_text(size = 7.5)) +
  guides(fill = guide_legend(nrow = 2))
```

The single largest proposed item is **By-Pass & Diversion Roads (₱127B)** — a
national Network Development line that Congress has historically *cut* by the most
(recall the ₱94B cumulative reduction above). Right behind it is **BIP – Access
Roads to Public Buildings (₱106B)**, a Convergence item of exactly the type
Congress historically *adds*. The two big flood-mitigation lines return at ₱63B and
₱41B. In other words, the FY 2027 NEP re-proposes a large national road-and-flood
program — much of which the legislature has a track record of trimming in favour
of district items — setting up the same proposal-versus-enactment tug-of-war at
the item level that the rest of this report documents at the program and district
levels.

## By District Engineering Office: The Proposed Distribution

Having seen how Congress has historically built up district budgets, it is worth
turning the same geographic lens on the **FY 2027 proposal**. Its starting
distribution across operating units is strikingly centralized — a reminder that
the district-level build-up documented above happens *after* the proposal, not in
it.

Of the ₱642.6B proposed for FY 2027, **₱510.0B (79%) sits in the Central Office**
and only **₱129.7B (20%) is distributed across the 198 district engineering
offices** (the small remainder is in regional and other units). That 79/20 split
is the raw material Congress works on: the district-level build-up that defines
DPWH's budget politics happens *later*, in the move from proposal to enacted law.

```{r plot-levels-centraldist, fig.height=3.8}
unit_levels <- byoperunit %>%
  mutate(Unit = case_when(Office == "Central Office"        ~ "Central Office",
                          str_detect(Office, "D.E. Office") ~ "District Engineering Offices",
                          str_detect(Office, "Regional")    ~ "Regional Offices",
                          TRUE                              ~ "Other units")) %>%
  filter(FY >= 2018) %>%
  group_by(FY, Unit) %>%
  summarise(NEP_B = sum(NEP_B, na.rm = TRUE), .groups = "drop") %>%
  group_by(FY) %>% mutate(share = NEP_B / sum(NEP_B)) %>% ungroup() %>%
  mutate(Unit = factor(Unit, levels = c("Central Office", "District Engineering Offices",
                                         "Regional Offices", "Other units")))

ggplot(unit_levels, aes(x = FY, y = share, fill = Unit)) +
  geom_col(width = 0.74) +
  scale_x_continuous(breaks = c(seq(2018, 2026, 2), 2027)) +
  scale_y_continuous(labels = scales::percent, expand = expansion(mult = c(0, 0.02))) +
  scale_fill_manual(values = c("Central Office" = "#3994fc",
                               "District Engineering Offices" = "#ffa6b0",
                               "Regional Offices" = "#bcbef2", "Other units" = "#d9d9d9")) +
  labs(x = NULL, y = "Share of proposed NEP", fill = NULL,
       title = "Where the Proposal Places the Money: NEP Share by Operating Unit",
       caption = "**Note:** Each bar is the FY's *proposed* (NEP) new appropriations, split by operating unit. FY2027 is proposed only.<br>**Visualization:** AJ Montesa / People's Budget Coalition.") +
  theme_bw() +
  theme(legend.position = "bottom", legend.text = element_text(size = 7),
        panel.grid.minor = element_blank(), panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold", size = 12.5),
        plot.caption = element_markdown(size = 6)) +
  guides(fill = guide_legend(nrow = 1))
```

At proposal stage the Central Office has held the overwhelming share of new
appropriations every year — typically 75-85% — because the Executive budgets
most district projects centrally and lets the release-and-realignment machinery
distribute them. The FY2027 proposal (79% central) is squarely in that pattern.

**Which districts fare best in the proposal?** Even within the 20% that is
pre-allocated to named districts, the FY2027 NEP is concentrated: the top office
is proposed ₱3.5B while the median district gets a fraction of that.

```{r plot-levels-topdeo, fig.height=5.0}
top_deo_27 <- deo_all %>%
  filter(FY == 2027) %>%
  slice_max(NEP_B, n = 20) %>%
  mutate(Office = str_remove(Office, " D\\.E\\. Office$") %>% str_remove(" Engineering$"),
         Office = fct_reorder(Office, NEP_B))

ggplot(top_deo_27, aes(x = NEP_B, y = Office)) +
  geom_col(fill = "#3994fc", width = 0.72) +
  geom_text(aes(label = format(round(NEP_B, 2), nsmall = 2)), hjust = -0.15, size = 2.6) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = "FY 2027 proposed NEP (PHP billions)", y = NULL,
       title = "Top 20 District Engineering Offices in the FY 2027 Proposal",
       caption = "**Note:** Pre-allocated district amounts in the NEP, before congressional realignment.<br>**Visualization:** AJ Montesa / People's Budget Coalition.") +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold", size = 12.5),
        plot.caption = element_markdown(size = 6),
        axis.text.y = element_text(size = 7))
```

The offices proposed the most in FY2027 — Bulacan 2nd (₱3.5B), Metro Manila 3rd
(₱2.9B), Nueva Ecija 1st (₱2.8B) — are a mix of high-population and
infrastructure-heavy districts. But the more important point for scrutiny is the
gap between this proposed geography and the *enacted* one: as the realignment
analysis above shows, Congress adds heavily to district offices during
deliberations, so the enacted distribution of DPWH money will look very different
from this proposal.


# The Line-Item Vetoes: 2019 and 2025

Twice in this window the President reached past Congress's realignments and struck
enacted amounts out of the DPWH budget by **line-item veto** — in **FY2019
(₱95.4B)** and again in **FY2025 (₱26.1B)**. In both years the pattern was the
same: Congress inflated the proposal (heavily in 2019's Local Program, and in
2025's Convergence and Flood programs), and the President then vetoed a large
block of it, almost all from the **Central Office**. The enacted GAA used
throughout this report is the **post-veto** figure in both years.

## The 2019 Veto (₱95.4B)

In FY2019, Congress raised the DPWH's New Appropriations only modestly in
aggregate (from a **₱544.5B** proposal to a **₱549.4B** legislated total) but **heavily
reallocated** the money — above all into the Local Program. The President then
exercised **line-item vetoes worth ₱95.37B**, almost all of it from that program,
bringing the enacted GAA down to **₱454.0B** (the NEP-to-GAA dip visible back in
Section 1). The three stages below are what the President **proposed** (NEP), what
Congress **legislated** (the pre-veto total), and what was finally **enacted**
after the line-item veto.

```{r plot-veto-total, fig.height=3.6}
veto_total <- tibble(
  Stage = factor(c("Proposed\n(NEP)", "Legislated\n(pre-veto)", "Enacted\n(post-veto)"),
                 levels = c("Proposed\n(NEP)", "Legislated\n(pre-veto)", "Enacted\n(post-veto)")),
  AmountB = c(sum(byprogram$NEP_B[byprogram$FY == 2019]),
              sum(byprogram$GAA_B[byprogram$FY == 2019]) + sum(byprogram$Veto_B[byprogram$FY == 2019]),
              sum(byprogram$GAA_B[byprogram$FY == 2019]))
)

ggplot(veto_total, aes(x = Stage, y = AmountB, fill = Stage)) +
  geom_col(width = 0.62) +
  geom_text(aes(label = comma(AmountB, accuracy = 0.1)), vjust = -0.5, size = 3) +
  annotate("segment", x = 2, xend = 3, y = 575, yend = 575) +
  annotate("text", x = 2.5, y = 590, label = "vetoed: -P95.4B", size = 3, fontface = "bold") +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.18))) +
  scale_fill_manual(values = c("#ffa6b0", "#bcbef2", "#3994fc")) +
  labs(x = NULL, y = "PHP billions",
       title = "DPWH FY2019: Proposed -> Legislated -> Enacted (after the Veto)") +
  theme_minimal() +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(), panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", size = 13),
        axis.text.x = element_text(size = 8))
```

```{r plot-veto-byprogram, fig.height=4.6}
veto_prog <- byprogram %>%
  filter(FY == 2019) %>%
  transmute(Program,
            `Proposed (NEP)`        = NEP_B,
            `Legislated (pre-veto)` = GAA_B + Veto_B,
            `Enacted (post-veto)`   = GAA_B) %>%
  pivot_longer(-Program, names_to = "Stage", values_to = "AmountB") %>%
  mutate(Stage   = factor(Stage, levels = c("Proposed (NEP)", "Legislated (pre-veto)", "Enacted (post-veto)")),
         Program = factor(Program, levels = dpwh_programs))

ggplot(veto_prog, aes(x = Stage, y = AmountB, fill = Stage)) +
  geom_col(width = 0.7) +
  facet_wrap(~Program, scales = "free_y", ncol = 4, labeller = label_wrap_gen(22)) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.1))) +
  scale_fill_manual(values = c("Proposed (NEP)" = "#ffa6b0", "Legislated (pre-veto)" = "#bcbef2",
                               "Enacted (post-veto)" = "#3994fc")) +
  labs(x = NULL, y = "PHP billions", fill = NULL,
       title = "FY2019 by Program: Proposed, Legislated, and Enacted after the Veto",
       caption = "**Data Sources:** NEP, legislated GAA, and veto message, FY2019.<br>**Visualization:** AJ Montesa / People's Budget Coalition.") +
  theme_bw() +
  theme(legend.position = "bottom", legend.text = element_text(size = 7),
        panel.grid.minor = element_blank(), panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold", size = 13),
        plot.caption = element_markdown(size = 6),
        strip.background = element_rect(fill = "white"), strip.text = element_text(size = 6.5),
        axis.text.x = element_text(angle = 35, hjust = 1, size = 6),
        axis.text.y = element_text(size = 6.5))
```

The veto was overwhelmingly a **Local Program** event: only that program (plus
token amounts in the two overhead lines) shows a legislated (pre-veto) bar towering over the
enacted level. Unpacking the vetoed line items by **type of work** and by **engineering
office** shows where the realigned money had been parked:

```{r load-veto-detail}
veto_detail <- read_xlsx(data_path, sheet = "2019_Veto") %>%
  mutate(Veto_B = GAA_Veto / 1e9)
```

```{r plot-veto-activity, fig.height=3.8}
veto_detail %>%
  group_by(Activity) %>%
  summarize(Veto_B = sum(Veto_B), .groups = "drop") %>%
  ggplot(aes(x = Veto_B, y = fct_reorder(Activity, Veto_B))) +
  geom_col(fill = "#3994fc") +
  geom_text(aes(label = format(round(Veto_B, 1), nsmall = 1)), hjust = -0.15, size = 2.6) +
  scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.12))) +
  labs(x = "PHP billions vetoed", y = NULL,
       title = "What the 2019 Veto Struck, by Type of Work",
       caption = "**Data Source:** FY2019 veto message.<br>**Visualization:** AJ Montesa / People's Budget Coalition.") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(), panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", size = 13),
        plot.caption = element_markdown(size = 6),
        axis.text.y = element_text(size = 8))
```

```{r plot-veto-office, fig.height=4.4}
veto_detail %>%
  group_by(Office) %>%
  summarize(Veto_B = sum(Veto_B), .groups = "drop") %>%
  slice_max(Veto_B, n = 15) %>%
  ggplot(aes(x = Veto_B, y = fct_reorder(Office, Veto_B))) +
  geom_col(fill = "#ffa6b0") +
  geom_text(aes(label = format(round(Veto_B, 2), nsmall = 2)), hjust = -0.15, size = 2.4) +
  scale_x_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.14))) +
  labs(x = "PHP billions vetoed", y = NULL,
       title = "Where the 2019 Veto Struck: Top 15 Offices",
       subtitle = "The DPWH Central Office alone accounts for ~P74B of the P95.4B vetoed",
       caption = "**Data Source:** FY2019 veto message.<br>**Visualization:** AJ Montesa / People's Budget Coalition.") +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(), panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(size = 9),
        plot.caption = element_markdown(size = 6),
        axis.text.y = element_text(size = 7.5))
```

```{r table-veto-program}
byprogram %>%
  filter(FY == 2019) %>%
  transmute(Program,
            `Proposed (NEP)`        = NEP_B,
            `Legislated (pre-veto)` = GAA_B + Veto_B,
            `Vetoed`                = Veto_B,
            `Enacted (post-veto)`   = GAA_B) %>%
  bind_rows(tibble(Program = "DPWH TOTAL",
                   `Proposed (NEP)` = sum(byprogram$NEP_B[byprogram$FY == 2019]),
                   `Legislated (pre-veto)` = sum((byprogram$GAA_B + byprogram$Veto_B)[byprogram$FY == 2019]),
                   Vetoed = sum(byprogram$Veto_B[byprogram$FY == 2019]),
                   `Enacted (post-veto)` = sum(byprogram$GAA_B[byprogram$FY == 2019]))) %>%
  mutate(across(-Program, ~comma(.x, accuracy = 0.1))) %>%
  kbl(align = c("l", rep("r", 4)),
      caption = "FY2019 DPWH: proposed, legislated (pre-veto), vetoed, and enacted (post-veto) --- PHP billions") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 12) %>%
  row_spec(0, bold = TRUE)
```

```{r table-veto-activity}
veto_detail %>%
  group_by(Activity) %>%
  summarize(`Vetoed (PHP B)` = sum(Veto_B), .groups = "drop") %>%
  arrange(desc(`Vetoed (PHP B)`)) %>%
  mutate(`Share of veto` = percent(`Vetoed (PHP B)` / sum(`Vetoed (PHP B)`), accuracy = 0.1),
         `Vetoed (PHP B)` = comma(`Vetoed (PHP B)`, accuracy = 0.01)) %>%
  kbl(align = "lrr",
      caption = "FY2019 vetoed line items, by type of work") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 12)
```

## The 2025 Veto (₱26.1B)

FY2025 repeated the pattern on a smaller scale. Congress raised the DPWH proposal
sharply — from a **₱898.9B** NEP to a **₱1,113.8B** legislated total, a **+₱215B**
insertion — and the President then vetoed **₱26.1B**, bringing the enacted GAA
to **₱1,087.7B**. Where 2019's veto fell on the Local Program, 2025's landed
squarely on the two programs most associated with discretionary spending:
**Convergence and Special Support (₱18.0B vetoed)** and **Flood Management
(₱7.5B)**.

```{r plot-veto2025-total, fig.height=3.6}
veto_total_25 <- tibble(
  Stage = factor(c("Proposed\n(NEP)", "Legislated\n(pre-veto)", "Enacted\n(post-veto)"),
                 levels = c("Proposed\n(NEP)", "Legislated\n(pre-veto)", "Enacted\n(post-veto)")),
  AmountB = c(sum(byprogram$NEP_B[byprogram$FY == 2025]),
              sum(byprogram$GAA_B[byprogram$FY == 2025]) + sum(byprogram$Veto_B[byprogram$FY == 2025]),
              sum(byprogram$GAA_B[byprogram$FY == 2025]))
)

ggplot(veto_total_25, aes(x = Stage, y = AmountB, fill = Stage)) +
  geom_col(width = 0.62) +
  geom_text(aes(label = comma(AmountB, accuracy = 0.1)), vjust = -0.5, size = 3) +
  annotate("segment", x = 2, xend = 3, y = 1150, yend = 1150) +
  annotate("text", x = 2.5, y = 1175, label = "vetoed: -P26.1B", size = 3, fontface = "bold") +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.18))) +
  scale_fill_manual(values = c("#ffa6b0", "#bcbef2", "#3994fc")) +
  labs(x = NULL, y = "PHP billions",
       title = "DPWH FY2025: Proposed -> Legislated -> Enacted (after the Veto)") +
  theme_minimal() +
  theme(legend.position = "none",
        panel.grid.major.x = element_blank(), panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold", size = 13),
        axis.text.x = element_text(size = 8))
```

The vetoed amounts fell on three programs, and — as in 2019 — almost entirely
on centrally-held appropriations:

```{r plot-veto2025-detail, fig.height=3.4}
veto25_prog <- byprogram %>%
  filter(FY == 2025, Veto_B > 0) %>%
  transmute(Program = factor(Program, levels = dpwh_programs), Veto_B) %>%
  arrange(desc(Veto_B))

central_share_25 <- read_xlsx(data_path, sheet = "2025_Veto") %>%
  summarise(sh = sum(GAA_Veto[Office == "Central Office"], na.rm = TRUE) /
                 sum(GAA_Veto, na.rm = TRUE)) %>% pull(sh)

ggplot(veto25_prog, aes(x = Veto_B, y = fct_reorder(Program, Veto_B))) +
  geom_col(fill = "#8c8c8c", width = 0.6) +
  geom_text(aes(label = comma(Veto_B, accuracy = 0.1)), hjust = -0.15, size = 3) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.15))) +
  scale_y_discrete(labels = function(x) str_wrap(x, 24)) +
  labs(x = "Vetoed (PHP billions)", y = NULL,
       title = "FY2025 Line-Item Veto, by Program (₱26.1B total)",
       subtitle = sprintf("The Central Office accounts for ~₱25.1B (%.0f%%) of the amount vetoed",
                          100 * central_share_25),
       caption = "**Data Source:** FY2025 veto message.<br>**Visualization:** AJ Montesa / People's Budget Coalition.") +
  theme_bw() +
  theme(panel.grid.minor = element_blank(), panel.grid.major.y = element_blank(),
        plot.title = element_text(face = "bold", size = 12.5),
        plot.subtitle = element_text(size = 9), plot.caption = element_markdown(size = 6),
        axis.text.y = element_text(size = 8))
```

As in 2019, the veto was overwhelmingly a **Central Office** event — about
**₱25.1B (96%)** of the ₱26.1B struck was drawn from centrally-held appropriations,
with only token amounts removed from named district offices. The recurring story
is the same across both years: the largest discretionary build-ups are inserted
centrally during deliberations, and it is there that the veto pen, when used,
falls.


# DPWH and the Unprogrammed Appropriations

The **Unprogrammed Appropriations (UA)** is a standby fund that may only be
released when revenue collections exceed targets or new financing is secured.
Because releases happen *outside* the regular program structure, they are harder
to track — and the DPWH's draw on the UA has become a focus of scrutiny, as a
number of flood-control projects later flagged as questionable were reportedly
financed through this route. The table records **allotments released** (the
spending authority granted) and **obligations** (amounts actually committed);
obligation data before 2013 could not be sourced, and FY2015-2017 carried no UA
releases to the DPWH at all.

```{r load-ua}
ua <- read_xlsx(data_path, sheet = "DPWH_UA") %>%
  mutate(Allotments_B  = Allotments / 1e6,    # thousand pesos -> PHP billions
         Obligations_B = Obligations / 1e6)

ua_long <- ua %>%
  select(FY, Allotments_B, Obligations_B) %>%
  pivot_longer(-FY, names_to = "Item", values_to = "AmountB") %>%
  mutate(Item = recode(Item, Allotments_B = "Allotments", Obligations_B = "Obligations"),
         Item = factor(Item, levels = c("Allotments", "Obligations")))
```

```{r plot-ua, fig.height=4.2}
ggplot(ua_long %>% filter(!is.na(AmountB)), aes(x = FY, y = AmountB, fill = Item)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.66) +
  scale_x_continuous(breaks = 2011:2025) +
  scale_y_continuous(labels = scales::comma, expand = expansion(mult = c(0, 0.06))) +
  scale_fill_manual(values = c("Allotments" = "#ffa6b0", "Obligations" = "#3994fc")) +
  labs(x = NULL, y = "PHP billions", fill = NULL,
       title = "DPWH Releases from the Unprogrammed Appropriations",
       subtitle = "Allotments released and obligations incurred, by fiscal year",
       caption = "**Data Sources:** DBM allotment and obligation records, 2011-2025.<br>**Visualization:** AJ Montesa / People's Budget Coalition.") +
  theme_minimal() +
  theme(legend.position = c(0.15, 0.75),
        panel.grid.minor = element_blank(), panel.grid.major.x = element_blank(),
        plot.title = element_text(face = "bold", size = 13),
        plot.subtitle = element_text(size = 9),
        plot.caption = element_markdown(size = 6),
        axis.text.x = element_text(size = 7))
```

After a decade of negligible draws, UA releases to the DPWH surged from **₱16B in
2022 to ₱89B in 2023 and ₱245B in 2024** — the latter alone roughly a quarter of
the department's entire regular budget that year. Yet a large share of these
allotments went unobligated (about **54% utilised in 2024**, **50% in 2025**),
which — alongside the project-level findings reported elsewhere — is exactly the
kind of pattern that merits questioning at the hearings.

```{r table-ua}
ua %>%
  transmute(FY,
            `Allotments (B)`  = Allotments_B,
            `Obligations (B)` = Obligations_B,
            Utilization       = Obligations_B / Allotments_B) %>%
  mutate(
    `Allotments (B)`  = comma(`Allotments (B)`, accuracy = 0.01),
    `Obligations (B)` = ifelse(is.na(`Obligations (B)`), "\u2013",
                               comma(`Obligations (B)`, accuracy = 0.01)),
    Utilization       = ifelse(is.na(Utilization) | is.nan(Utilization) | is.infinite(Utilization),
                               "\u2013", percent(Utilization, accuracy = 1))
  ) %>%
  kbl(align = "lrrr",
      caption = "DPWH releases from the Unprogrammed Appropriations, PHP billions") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE, font_size = 12)
```

# Notes

**On coverage.** The Section 1 series is intentionally left with gaps where the
physical documents have not yet been encoded; points mark observed years. The
by-operating-unit data now runs from the FY2014 budget through FY2026, so the
department-wide comparisons span the 16th through 20th Congresses and seven House
Speakers (see the Speaker table in the operating-unit section). The district-office
deep-dives center on the Duterte and Marcos administrations (FY2017 onward —
ten budgets, 2019 included), though the realignments-by-Congress chart reaches back
to the 16th Congress (the three Aquino-era budgets) for context. The FY2019 figures
use the enacted, post-veto GAA; the ₱95.4B line-item veto that year is isolated in
its own chart and section.

**On the 2019 GAA.** The GAA figures throughout are the **enacted, post-veto**
levels (they reconcile to the Office-of-the-Secretary totals in Section 1).
"Legislated (pre-veto)" reconstructs what Congress passed by adding the vetoed
amounts back to the enacted GAA. Because of this, every 2019 realignment figure in
this document reflects the *combined* effect of Congress's reallocation and the
President's veto, not Congress alone; the veto section, the Central-vs-District veto
chart, and the grey segments in the realignments-by-Congress chart isolate the veto
and recover Congress's pre-veto intent.

**On the Unprogrammed Appropriations.** UA obligation data before 2013 could not
be sourced; FY2015-2017 had no UA releases to the DPWH. These realignments are simply the arithmetic NEP-to-GAA difference and are not, on
their own, evidence of irregularity.
